---
title: "Co-occurence pre-processing"
author: "Emily"
date: "2025-05-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(sf)
library(terra)
library(rnaturalearth)
```

# Site pre-processing

## Import raw station layout data files 
                                                
```{r}
# read Ruaha-Rungwa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Ruaha-Rungwa_CTtrap_layouts")

RRlayoutMIO2018 <- read.csv("TZ_RR_MIO-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutRNP2018 <- read.csv("TZ_RR_RNP-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutWMA2018 <- read.csv("TZ_RR_WMA-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutRGI2019 <- read.csv("TZ_RR_RGI-2019_CT trap layout_midday.csv", header = FALSE)

# read Selous-Nyerere data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Selous-Nyere_CTtrap_layouts")

SNlayoutNMTL2020 <- read.csv("TZ_SN_NMTL-2020_CT trap layout_midday.csv", header = FALSE)
SNlayoutSKPE2021 <- read.csv("TZ_SN_SKPE-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSKPW2021 <- read.csv("TZ_SN_SKPW-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSLWN2021 <- read.csv("TZ_SN_SLWN-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSMGN2021 <- read.csv("TZ_SN_SMGN-2021_CT trap layout_midday_excl missing data.csv", header = FALSE)
SNlayoutNKLC2022 <- read.csv("TZ_SN_NKLC-2022_CT trap layout_midday.csv", header = FALSE)
SNlayoutNMSK2022 <- read.csv("TZ_SN_NMSK-2022_CT trap layout_midday.csv", header = FALSE)

# read Lower Luangwa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Lower_Luangwa_CTtrap_layouts")

ZBlayoutLL2020 <- read.csv("ZB-LU-2020_CT layout midday.csv", header = TRUE)
ZBlayoutLL2021 <- read.csv("ZB-LU-2021_CT layout midday.csv", header = TRUE)
ZBlayoutLL2022 <- read.csv("ZB-LU-2022_CT layout midday.csv", header = TRUE)

# read Rufunsa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Rufunsa_CTtrap_layouts")

ZBlayoutR2020 <- read.csv("ZB-RUF-2020_CT layout midday.csv", header = TRUE)
ZBlayoutR2021 <- read.csv("ZB-RUF-2021_CT layout midday.csv", header = TRUE)
ZBlayoutR2022 <- read.csv("ZB-RUF-2022_CT layout midday.csv", header = TRUE)
```

## Create spatial dataframes

### Tanzania

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022") 

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 1:3] # Keep only the first 3 columns LOC_ID, X, Y
  obj <- obj[-c(1, 2), ] # Remove the first two rows - they are blank
  colnames(obj) <- as.character(obj[1, ]) # Set new row 1 as column names
  obj <- obj[-1, ] # Remove the new first row which repeats the column names
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c( "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial", "RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial")

# Loop through all datasets and remove rows with missing coordinates - this is just removing blank rows that shouldn't be there
for (ds in datasets) {
  df <- get(ds)
  df_clean <- df[!(is.na(df$X_UTM37S) | is.na(df$Y_UTM37S) |
                   df$X_UTM37S == "" | df$Y_UTM37S == ""), ] # Remove rows where X or Y is NA or blank
  assign(ds, df_clean, envir = .GlobalEnv) # Assign cleaned data back to the environment
}

# Project into Africa Albers
# Define Africa Albers Equal Area CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# UMT37S datasets
datasets <- c( "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial")

# Define UTM Zone 37S CRS (EPSG:32737)
utm_37s_crs <- st_crs(32737)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X_UTM37S", "Y_UTM37S"), crs = utm_37s_crs) # Convert to sf object with UTM coordinates
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinates
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}

# List UTM36S datasets
datasets <- c("RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial")

# UTM Zone 36S CRS
utm_36s_crs <- st_crs(32736)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X_UTM37S", "Y_UTM37S"), crs = utm_36s_crs) # Convert to sf object using the coordinates (still named as 37S in columns, but actually 36S)
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinate
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}

```

### Zambia

```{r}
datasets <- c("ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022")

# Change column name to LOC_ID and remove Cam ID column, collapse to per station not camera
for (name in datasets) {
  df <- get(name)
  colnames(df)[1] <- "LOC_ID"  # Rename column to LOC_ID
  df <- df[, -2]  # Remove the second column (Cam ID)
  date_cols <- setdiff(names(df), c("LOC_ID", "Y", "X")) # Identify date columns
  df <- df %>%
    group_by(LOC_ID) %>%
    summarise(
      Y = first(Y), # Take first X coord (they should both be the same)
      X = first(X), # Take first y coord (they should both be the same)
      across(all_of(date_cols), ~ if (all(is.na(.))) NA else as.integer(any(. == 1, na.rm = TRUE))), .groups = "drop") # If one of the values is 1, print 1, if anything else print NA
  df[is.na(df)] <- 0 # Change NAs to 0s to show station not operational
  colnames(df)[-3] <- sub("^X", "", colnames(df)[-3]) # Remove X from dates (they currently have an X at the start of them)
  assign(name, df, envir = .GlobalEnv) # Save back over original file
}

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 1:3] # Keep only the first 3 columns (LOC_ID, X, Y)
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c("ZBlayoutLL2020_spatial", "ZBlayoutLL2021_spatial", "ZBlayoutLL2022_spatial", "ZBlayoutR2020_spatial",  "ZBlayoutR2021_spatial", "ZBlayoutR2022_spatial")

# Loop through and swap round Y and X columns for projection
for (ds in datasets) {
  df <- get(ds)  # Get the datasets
  df <- df[, c(1, 3, 2)]  # Reorder columns: first column, then third, then second
  assign(ds, df, envir = .GlobalEnv) # Overwrite the original dataset with the modified one
}

# Project into Africa Albers
# Define Africa Albers Equal Area CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# Define WGS84 CRS for GPS coordinates
wgs84_crs <- st_crs(4326)

# Loop through each dataset in the list
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X", "Y"), crs = wgs84_crs)  # Convert to sf object with decimal degrees (WGS84) coordinate
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinates (in meters in Africa Albers)
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}
```

### Save spatial datasets to drive

```{r}
datasets <- c("RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial", "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial", "ZBlayoutLL2020_spatial", "ZBlayoutLL2021_spatial", "ZBlayoutLL2022_spatial", "ZBlayoutR2020_spatial",  "ZBlayoutR2021_spatial", "ZBlayoutR2022_spatial")

# Define output directory
output_dir <- "C:/Users/lady7240.BIOLOGY/OneDrive - Nexus365/Documents/Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all datasets
for (ds in datasets) {
  file_name <- paste0(ds, ".csv")
  file_path <- file.path(output_dir, file_name)
  write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

## Prepare layout files for detection hitories  

### Tanzania

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022")

# Loop through to remove location data and end day date
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[-c(2,3), -c(2, 3)] # Remove row 2 and columns 2 & 3
  obj[1, ] <- sub(" .*", "", obj[1, ]) # Keep only date part in row 1
  colnames(obj) <- as.character(obj[1, ]) # Set row 1 as column names
  obj <- obj[-1, ] # Remove the new first row
  assign(ds, obj, envir = .GlobalEnv) # Assign cleaned dataset back to the environment
}
```

### Zambia

```{r}
# NOTE: Relies on cleaned dataset from spatial layer creation
datasets <- c("ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022")

# Loop through and modify each dataset to remove location data and match date format to Tanzania: DD.MM.YYYY -> DD/MM/YYYY and remove name of first column LOC_ID
for (ds in datasets) {
  df <- get(ds)
  df <- df[-c(2,3), -c(2, 3)] # Remove row 2 and columns 2 & 3
  new_names <- colnames(df) 
  date_cols <- grepl("^\\d{2}\\.\\d{2}\\.\\d{4}$" , new_names)# Identify which column names match the date pattern
  new_names[date_cols] <- gsub("\\.", "/", new_names[date_cols]) # Replace dots with slashes in those column names
  colnames(df) <- new_names  # Apply updated names and overwrite original dataset
  names(df)[1] <- ""  # Remove name of first column
  assign(ds, df, envir = .GlobalEnv)
}

#ZBlayoutLL2020 <- ZBlayoutLL2020[, -98]
```

### Save clean layout datasets

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022","ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022")

# Define output directory
output_dir <- "C:/Users/lady7240.BIOLOGY/OneDrive - Nexus365/Documents/Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all objects in the environment
for (ds in datasets) {
    file_name <- paste0(ds, "_clean.csv")
    file_path <- file.path(output_dir, file_name)
    write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

# Species pre-processing

## Import raw detection data files
```{r}
# read Ruaha-Rungwa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Ruaha-Rungwa_CTcapture_data")

RRdataMIO2018 <- read.csv("TZ_RR_MIO-2018_CT captures_202408.csv", header = FALSE)
RRdataRNP2018 <- read.csv("TZ_RR_RNP-2018_CT captures_202408.csv", header = FALSE)
RRdataWMA2018 <- read.csv("TZ_RR_WMA-2018_CT captures_202408.csv", header = FALSE)
RRdataRGI2019 <- read.csv("TZ_RR_RGI-2019_CT captures_202408.csv", header = FALSE)

# read Selous-Nyerere data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Selous-Nyere_CTcapture_data")

SNdataNMTL2020 <- read.csv("TZ_SN_NMTL-2020_CT captures_202405.csv", header = FALSE)
SNdataSKPE2021 <- read.csv("TZ_SN_SKPE-2021_CT captures_202405.csv", header = FALSE)
SNdataSKPW2021 <- read.csv("TZ_SN_SKPW-2021_CT captures_202405.csv", header = FALSE)
SNdataSLWN2021 <- read.csv("TZ_SN_SLWN-2021_CT captures_202405.csv", header = FALSE)
SNdataSMGN2021 <- read.csv("TZ_SN_SMGN-2021_CT captures_202405.csv", header = FALSE)
SNdataNKLC2022 <- read.csv("TZ_SN_NKLC-2022_CT captures_202405.csv", header = FALSE)
SNdataNMSK2022 <- read.csv("TZ_SN_NMSK-2022_CT captures_202405.csv", header = FALSE)

# read Lower Luangwa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Lower_Luangwa_CTcapture_data")

ZBdataLL2020 <- read.csv("All_Species_2020.csv", header = TRUE)
ZBdataLL2021 <- read.csv("All_Species_2021.csv", header = TRUE)
ZBdataLL2022 <- read.csv("LLZ_BM_CT_LowerLuangwa_2022.csv", header = TRUE)

# read Rufunsa data
setwd("~/Thesis/Chapter 3 - Co-occcurence/Data/Rufunsa_CTcapture_data")

ZBdataR2020 <- read.csv("All_Species_2020.csv", header = TRUE)
ZBdataR2021 <- read.csv("All_Species_2021.csv", header = TRUE)
ZBdataR2022 <- read.csv("All_Species_2022.csv", header = TRUE)
```

## Prepare detection data as lists and count detections per site x day x species

### Tanzania

```{r}
# List of datasets
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022")

# Loop over the datasets and set second row as column names and convert DATE to Date class
for (ds in datasets) {
  df <- get(ds) # Access the dataset using get() to modify it
  df <- df[-1, ] # Remove the first row
  colnames(df) <- df[1, ] # Set the second row as column names
  df <- df[-1, ] # Remove the second row (now the header)
  df$DATE <- as.Date(df$DATE, format = "%d/%m/%Y") # Convert DATE column to Date class
  assign(ds, df) # Assign the cleaned data back to the original dataset name
}

# Count species detections per species per day per site
species <- c("lion", "cheetah", "leopard", "spotted hyaena", "striped hyaena", 
             "wild dog", "caracal", "serval", "civet", "honey badger", 
             "wildcat", "bat-eared fox", "aardwolf", "b-b jackal", "s-s jackal")
             
# Initialize an empty list to store the results
result_countsTZ <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    df <- get(ds) # get the dataset and use [[ to access the column
    filtered_data <- df[df[[sp]] == 1, ] # Filter the dataset for rows where the species column has a value of 1
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Get the unique combinations of DATE and LOC_ID
    total_unique_combinations <- nrow(unique_combinations) # Count the total number of unique combinations
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create a result name: dataset name_species
    result_countsTZ[[result_name]] <- total_unique_combinations # Store the result in the list with the result name
  }
}

# Extract 1 species to double check outputs (RRMIO2018 Lion - 46)
lion_data <- RRdataMIO2018[RRdataMIO2018$lion == 1, ] # Filter the dataset for rows where the lion column has a value of 1
unique_combinations <- unique(lion_data[, c("DATE", "LOC_ID")]) # Get the unique combinations of DATE and LOC_ID
total_unique_combinations <- nrow(unique_combinations) # Count the total number of unique combinations
total_unique_combinations # Print the result

# Turn into list of detentions rather than a matrix

# Loop through and subset each dataset to remove unwanted species and pivot into one row per detection
for (ds in datasets) {
  df <- get(ds)
  cols_to_keep <- c(1:8, which(colnames(df) %in% species)) # Get all matching columns: first 8 + any from species list
  cols_to_keep <- unique(cols_to_keep)  # Avoid duplicates
  df <- df[, cols_to_keep]
  species_cols <- intersect(colnames(df), species) # Identify columns that are in the species list
  df <- df[!apply(df[, species_cols], 1, function(row) all(as.numeric(row) == 0)), ] # Remove rows where all remaining species columns are 0
    df <- df %>%
    pivot_longer(
      cols = all_of(species_cols),
      names_to = "Species",
      values_to = "value"
    ) %>%
    filter(as.numeric(value) == 1) # Pivot longer and filter for presence == 1
  assign(ds, df, envir = .GlobalEnv)
}
```

### Zambia

#### Lower Luangwa  

```{r}
# NOTE: These are processed separately as there are and different issues for each file

# LL2020 cleaning
ZBdataLL2020 <- ZBdataLL2020 %>% rename(Species = Tags, DATE = Date, LOC_ID = Area) # Rename columns To = From
ZBdataLL2020 <- ZBdataLL2020 %>% mutate(Species = recode(Species,
    "African Wild Dog" = "Wild Dog - Lycaon pictus",
    "Wild Dog - Canis pictus" = "Wild Dog - Lycaon pictus",
    "Spotted Hyena - Crocuta crocuta" = "Spotted Hyaena - Crocuta crocuta"
  )) # Rename species
ZBdataLL2020$LOC_ID <- gsub("[A-Za-z]", "", ZBdataLL2020$LOC_ID) # Clean LOC_ID to match layout dataframe
ZBdataLL2020$DATE <- as.Date(ZBdataLL2020$DATE, format = "%d/%m/%Y") # Match date formats

# LL2021 cleaning
ZBdataLL2021 <- ZBdataLL2021 %>% rename(DATE = Date.and.Time, LOC_ID = Cell) # Rename columns
ZBdataLL2021 <- ZBdataLL2021 %>% # Pivot for multiple species columns into 1
  pivot_longer(
    cols = c(Species.1, Species.2, Species.3),
    names_to = "Species_Column",
    values_to = "Species"
  ) %>%
  filter(!is.na(Species) & Species != "" & Species != "NA")  # Remove rows where Species was NA
ZBdataLL2021$Species[ZBdataLL2021$Species == "Honeybadger - Mellivora capensis"] <-"Honey Badger - Mellivora capensis" # Rename species
ZBdataLL2021$DATE <- as.Date(ZBdataLL2021$DATE, format = "%d/%m/%Y") # Match date formats

# LL2022 cleaning
# Split the data into Photo, Date, Time, Tag
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(
    Photo = str_extract(PhotoDateTimeTag, "^.*(?=\\d{2}/\\d{2}/\\d{2})"), # Extract Photo (everything before the date
    DATE = str_extract(PhotoDateTimeTag, "\\d{2}/\\d{2}/\\d{2}"), # Extract Date (DD/MM/YY)
    Time = str_extract(PhotoDateTimeTag, "\\d{2}:\\d{2}:\\d{1,2}(?: [APM]{2})?"),# Extract Time (handles formats like HH:MM:S or HH:MM:SS, optional AM/PM)
    Species = str_extract(PhotoDateTimeTag, "(?<=\\d{2}:\\d{2}:\\d{1,2}(?: [APM]{2})?).*") %>% 
      str_trim() # Extract Species
  )
  
# Clean species names in LL2022 ~ do this pre splitting as it makes splitting easier
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = str_replace_all(Species, "\\d+_|LHS_|LS_|RS_|RHS_|_RHS|_LHS|Front\\s+end_|Back\\s+end_|Highlight_|Hghlight_|_Sub-adult|_Highlight|_White|White_|IR_|_IR|_Snare|Juvenile_|_Juvenile|2 indiv_|_2 indiv|_Wounded|Adult_|\\b\\d{1,2} [AP]M(?=\\w)|^_|\\s{2,}", ""), # Remove specific substrings from anywhere in the string | times eg."2 PM" or "11 AM" even when directly followed by a word | leading underscore | double spaces
  )

# Clean specific names pre-splitting
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = case_when(
    Species %in% c("African - Civettictis civetta_Civet", "Civet African - Civettictis civetta", "ivet African - Civettictis civetta", "Cvet African - Civettictis civetta") ~ "African Civet - Civettictis civetta",
    Species %in% c("Jackal_Side-Striped - Canis adustus", "Jackal Side-Striped - Canis adustus") ~ "Side-Striped Jackal - Canis adustus",
    Species %in% c("Hyaena_Spotted  - Crocuta crocuta", "Hyaena Spotted  - Crocuta crocuta", "Hyaena_Spotted - Crocuta crocuta") ~ "Spotted Hyaena - Crocuta crocuta",
    Species == "Honeybadger - Mellivora capensis" ~ "Honey Badger - Mellivora capensis",
    Species == "Wild cat African - Felis lybica" ~ "African Wildcat - Felis lybica",
    Species %in% c("dult_Wild Dog - Lycaon pictus", "uvenile_Wild Dog - Lycaon pictus", "ild Dog - Lycaon pictus", "Wld Dog - Lycaon pictus", "Back end_Wild Dog - Lycaon pictus") ~ "Wild Dog - Lycaon pictus",
    TRUE ~ Species
  ))

# Split the Species column for multiple species and pivot
ZBdataLL2022 <- ZBdataLL2022 %>%
  separate_wider_delim(Species, delim = "_", names_sep = "_part", too_few = "align_start") %>%
  pivot_longer(
    cols = c(Species_part1, Species_part2, Species_part3, Species_part4),
    names_to = "Species_Column",
    values_to = "Species"
  ) %>%
  filter(!is.na(Species) & Species != "" & Species != "NA")  # Remove rows where Species was NA

# Extra cleaning post splitting
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = case_when(
    Species %in% c("Hyaena Spotted- Crocuta crocuta", "Spotted- Crocuta crocuta") ~ "Spotted Hyaena - Crocuta crocuta",
    Species == "Civet African - Civettictis civetta" ~ "African Civet - Civettictis civetta",
    TRUE ~ Species
  ))

# Loc ID extraction - needs further cleaning
ZBdataLL2022 <- ZBdataLL2022 %>% mutate(
    LOC_ID = str_extract(Photo, "^.*?(?=R)"),              # Extract everything before 'R'
    LOC_ID = gsub("[A-Za-z_]", "", LOC_ID)                 # Remove letters and underscores
  )

# Match date formats
ZBdataLL2022$DATE <- as.Date(ZBdataLL2022$DATE, format = "%d/%m/%y")


                       # Count number of days with detections of each species per site

datasets <- c("ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022")
species <- c("Lion - Panthera leo", "cheetah", "Leopard - Panthera pardus", "Spotted Hyaena - Crocuta crocuta", "Wild Dog - Lycaon pictus", "Caracal - Caracal caracal", "Serval - Leptailurus serval", "African Civet - Civettictis civetta", "Honey Badger - Mellivora capensis", "African Wildcat - Felis lybica",  "Side-Striped Jackal - Canis adustus")

# Initialize an empty list to store the results
result_countsLL <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp) # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result name
    result_countsLL[[result_name]] <- total_unique_combinations # Store result
  }
}
```

#### Rufunsa

```{r}
# R2020 cleaning
ZBdataR2020 <- ZBdataR2020 %>% rename(Species = Species.1) # Rename column - no other species columns
ZBdataR2020$Species[ZBdataR2020$Species == "Spotted Hyena - Crocuta crocuta"] <-"Spotted Hyaena - Crocuta crocuta" #Rename Species
ZBdataR2020$DATE <- as.Date(ZBdataR2020$DATE, format = "%d/%m/%Y") # Match date format

# R2021 cleaning
ZBdataR2021 <- ZBdataR2021 %>% rename(Species = Species.1) # Other species column has no carnivores
ZBdataR2021$Species[ZBdataR2021$Species == "Honeybadger - Mellivora capensis"] <-"Honey Badger - Mellivora capensis" #Rename species
ZBdataR2021 <- ZBdataR2021 %>%
  mutate(
    DATETIME = as.POSIXct(DATETIME, format = "%d/%m/%Y", tz = "UTC"), 
    DATE = as.Date(DATETIME) # Extract and create DATE column
  )

# R2022 cleaning
ZBdataR2022 <- ZBdataR2022 %>% rename(Species = Species.1, Checkdate = DATE) # Rename columns
ZBdataR2022 <- ZBdataR2022 %>%
  mutate(
    DATETIME = as.POSIXct(DATETIME, format = "%d/%m/%Y", tz = "UTC"),  
    DATE = as.Date(DATETIME) # Extract and create DATE column
  )
ZBdataR2022 <- ZBdataR2022 %>%
  mutate(Species = recode(Species,
    "Honeybadger - Mellivora capensis" = "Honey Badger - Mellivora capensis",
    "Civet African - Civettictis civetta" = "African Civet - Civettictis civetta",
    "Jackal Side-Striped - Canis adustus" = "Side-Striped Jackal - Canis adustus",
    "Hyaena Spotted  - Crocuta crocuta" = "Spotted Hyaena - Crocuta crocuta"
  ))

# Count number of days with detections of each species per site
datasets <- c("ZBdataR2020", "ZBdataR2021", "ZBdataR2022")
species <- c("Lion - Panthera leo", "Leopard - Panthera pardus", "Spotted Hyaena - Crocuta crocuta", "Wild Dog - Lycaon pictus", "Caracal - Caracal caracal", "Serval - Leptailurus serval", "African Civet - Civettictis civetta", "Honey Badger - Mellivora capensis", "African Wildcat - Felis lybica", "Side-Striped Jackal - Canis adustus")

# Initialize an empty list to store the results
result_countsR <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp)  # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result nam
    result_countsR[[result_name]] <- total_unique_combinations # Store result
  }
}
```

### Match species names across datasets and remove unneccsary species data

```{r}
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022", "ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022", "ZBdataR2020", "ZBdataR2021", "ZBdataR2022")

# Create a named vector for recoding
species_recode <- c(
  "Lion - Panthera leo" = "Lion",
  "lion" = "Lion",
  "Leopard - Panthera pardus" = "Leopard",
  "leopard" = "Leopard",
  "cheetah" = "Cheetah",
  "Spotted Hyaena - Crocuta crocuta" = "Spotted hyaena",
  "spotted hyaena" = "Spotted hyaena",
  "striped hyaena" = "Striped hyaena",
  "Wild Dog - Lycaon pictus" = "African wild dog",
  "wild dog" = "African wild dog",
  "Caracal - Caracal caracal" = "Caracal",
  "caracal" = "Caracal",
  "Serval - Leptailurus serval" = "Serval",
  "serval" = "Serval",
  "African Civet - Civettictis civetta" = "African civet",
  "civet" = "African civet",
  "Honey Badger - Mellivora capensis" = "Honey badger",
  "honey badger" = "Honey badger",
  "African Wildcat - Felis lybica" = "African wildcat",
  "wildcat" = "African wildcat",
  "Side-Striped Jackal - Canis adustus" = "Side-striped jackal",
  "s-s jackal" = "Side-striped jackal",
  "bat-eared fox" = "Bat-eared fox",
  "aardwolf" = "Aardwolf",
  "b-b jackal" = "Black-backed jackal"
)

# Loop through all datasets and apply recoding
for (ds in datasets) {
  df <- get(ds)
  df <- df %>%
    mutate(Species = recode(Species, !!!species_recode)) %>% # Recode species
    filter(Species %in% unname(species_recode))  # Keep only recoded species
  assign(ds, df, envir = .GlobalEnv)
}
```

### Save detection datasets

```{r}
# Save clean datasets
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022", "ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022", "ZBdataR2020", "ZBdataR2021", "ZBdataR2022")

# Define output directory
output_dir <- "C:/Users/lady7240.BIOLOGY/OneDrive - Nexus365/Documents/Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all objects in the environment
for (ds in datasets) {
    file_name <- paste0(ds, "_clean.csv")
    file_path <- file.path(output_dir, file_name)
    write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

# Spatial covariate pre-processing

## Import camera trap locations and combine

```{r}
datasets <- ls()  
sp_list <- lapply(datasets[grepl("_spatial$", datasets)], get) # Select spatial dataframes
camlocs <- do.call(rbind, sp_list) # Merge 
camlocs_vec <- vect(camlocs, geom=c("X_AfricaAlbers", "Y_AfricaAlbers"), crs="+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25") # Convert to spatial file

# View data to sense check
mapview(camlocs_vec)
```

## Prepare landcover raster

```{r}
setwd("C:/Users/lady7240.BIOLOGY/OneDrive - Nexus365/Documents/Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Landcover/")

# List all CSV files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("lndc", i), rast(files[i]))
}

lndc_list <- sprc(lndc1, lndc2, lndc3, lndc4, lndc5, lndc6)
lndc <- mosaic(lndc_list)
plot(lndc)

# Project to Africa Albers
crs(lndc,describe=T) 

# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_contries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Mask landcover by countries 
```

## Prepare human pop raster
```{r}

```

## Prepare slope raster

## Prepare evelvation raster

## Prepare roughness raster

## Create buffer for covariate extraction

```{r}
# Create a buffer of 1000 m around the camera trap locations
cam_1000m_buffer <- buffer(camlocs_vec, width = 1000)
mapview(cam_1000m_buffer)
```

