---
title: "Co-occurence pre-processing"
author: "Emily"
date: "2025-05-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(sf)
library(terra)
library(rnaturalearth)
library(exactextractr)
library(raster)
library(ggplot2)
library(tidyterra)
library(mapview)
library(viridis)
library(spatstat)
library(gstat)
library(RANN)
library(corrplot)
```

# Site pre-processing

## Import raw station layout data files 
                                                
```{r}
# read Ruaha-Rungwa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Ruaha-Rungwa_CTtrap_layouts")

RRlayoutMIO2018 <- read.csv("TZ_RR_MIO-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutRNP2018 <- read.csv("TZ_RR_RNP-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutWMA2018 <- read.csv("TZ_RR_WMA-2018_CT trap layout_midday.csv", header = FALSE)
RRlayoutRGI2019 <- read.csv("TZ_RR_RGI-2019_CT trap layout_midday.csv", header = FALSE)

# read Selous-Nyerere data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Selous-Nyere_CTtrap_layouts")

SNlayoutNMTL2020 <- read.csv("TZ_SN_NMTL-2020_CT trap layout_midday.csv", header = FALSE)
SNlayoutSKPE2021 <- read.csv("TZ_SN_SKPE-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSKPW2021 <- read.csv("TZ_SN_SKPW-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSLWN2021 <- read.csv("TZ_SN_SLWN-2021_CT trap layout_midday.csv", header = FALSE)
SNlayoutSMGN2021 <- read.csv("TZ_SN_SMGN-2021_CT trap layout_midday_excl missing data.csv", header = FALSE)
SNlayoutNKLC2022 <- read.csv("TZ_SN_NKLC-2022_CT trap layout_midday.csv", header = FALSE)
SNlayoutNMSK2022 <- read.csv("TZ_SN_NMSK-2022_CT trap layout_midday.csv", header = FALSE)

# read Lower Luangwa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Lower_Luangwa_CTtrap_layouts")

ZBlayoutLL2020 <- read.csv("ZB-LU-2020_CT layout midday.csv", header = TRUE)
ZBlayoutLL2021 <- read.csv("ZB-LU-2021_CT layout midday.csv", header = TRUE)
ZBlayoutLL2022 <- read.csv("ZB-LU-2022_CT layout midday.csv", header = TRUE)

# read Rufunsa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Rufunsa_CTtrap_layouts")

ZBlayoutR2020 <- read.csv("ZB-RUF-2020_CT layout midday.csv", header = TRUE)
ZBlayoutR2021 <- read.csv("ZB-RUF-2021_CT layout midday.csv", header = TRUE)
ZBlayoutR2022 <- read.csv("ZB-RUF-2022_CT layout midday.csv", header = TRUE)

# read Laikipia data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Laikipia")
LKlayout2025 <- read.csv("KLKE_CT trap layout_midday.csv", header = TRUE)

# read Maasai Mara data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/BHP")
MMlayout2018 <- read.csv("all_sites_effort_occupancy_v2_2018.csv", header = TRUE)

# read Oljogi data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Ol Jogi")
OJlayout2022 <- read.csv("Oljogi.csv", header = TRUE)
```

## Create spatial dataframes

### Tanzania

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022") 

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 1:3] # Keep only the first 3 columns LOC_ID, X, Y
  obj <- obj[-c(1, 2), ] # Remove the first two rows - they are blank
  colnames(obj) <- as.character(obj[1, ]) # Set new row 1 as column names
  obj <- obj[-1, ] # Remove the new first row which repeats the column names
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c( "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial", "RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial")

# Loop through all datasets and remove rows with missing coordinates - this is just removing blank rows that shouldn't be there
for (ds in datasets) {
  df <- get(ds)
  df_clean <- df[!(is.na(df$X_UTM37S) | is.na(df$Y_UTM37S) |
                   df$X_UTM37S == "" | df$Y_UTM37S == ""), ] # Remove rows where X or Y is NA or blank
  assign(ds, df_clean, envir = .GlobalEnv) # Assign cleaned data back to the environment
}

# Project into Africa Albers
# Define Africa Albers Equal Area CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# UMT37S datasets
datasets <- c( "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial")

# Define UTM Zone 37S CRS (EPSG:32737)
utm_37s_crs <- st_crs(32737)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X_UTM37S", "Y_UTM37S"), crs = utm_37s_crs) # Convert to sf object with UTM coordinates
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinates
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}

# List UTM36S datasets
datasets <- c("RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial")

# UTM Zone 36S CRS
utm_36s_crs <- st_crs(32736)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X_UTM37S", "Y_UTM37S"), crs = utm_36s_crs) # Convert to sf object using the coordinates (still named as 37S in columns, but actually 36S)
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinate
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}
```

### Zambia

```{r}
datasets <- c("ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022")

# Change column name to LOC_ID and remove Cam ID column, collapse to per station not camera
for (name in datasets) {
  df <- get(name)
  colnames(df)[1] <- "LOC_ID"  # Rename column to LOC_ID
  df <- df[, -2]  # Remove the second column (Cam ID)
  date_cols <- setdiff(names(df), c("LOC_ID", "Y", "X")) # Identify date columns
  df <- df %>%
    group_by(LOC_ID) %>%
    summarise(
      Y = first(Y), # Take first X coord (they should both be the same)
      X = first(X), # Take first y coord (they should both be the same)
      across(all_of(date_cols), ~ if (all(is.na(.))) NA else as.integer(any(. == 1, na.rm = TRUE))), .groups = "drop") # If one of the values is 1, print 1, if anything else print NA
  df[is.na(df)] <- 0 # Change NAs to 0s to show station not operational
  colnames(df)[-3] <- sub("^X", "", colnames(df)[-3]) # Remove X from dates (they currently have an X at the start of them)
  assign(name, df, envir = .GlobalEnv) # Save back over original file
}

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 1:3] # Keep only the first 3 columns (LOC_ID, X, Y)
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c("ZBlayoutLL2020_spatial", "ZBlayoutLL2021_spatial", "ZBlayoutLL2022_spatial", "ZBlayoutR2020_spatial",  "ZBlayoutR2021_spatial", "ZBlayoutR2022_spatial")

# Loop through and swap round Y and X columns for projection
for (ds in datasets) {
  df <- get(ds)  # Get the datasets
  df <- df[, c(1, 3, 2)]  # Reorder columns: first column, then third, then second
  assign(ds, df, envir = .GlobalEnv) # Overwrite the original dataset with the modified one
}

# Append year onto loc-IDs due to location inconsistencies
ZBlayoutLL2020_spatial <- ZBlayoutLL2020_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2020"))
ZBlayoutLL2021_spatial <- ZBlayoutLL2021_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2021"))
ZBlayoutLL2022_spatial <- ZBlayoutLL2022_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2022"))
ZBlayoutR2020_spatial <- ZBlayoutR2020_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2020"))
ZBlayoutR2021_spatial <- ZBlayoutR2021_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2021"))
ZBlayoutR2022_spatial <- ZBlayoutR2022_spatial %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2022"))

# Project into Africa Albers
# Define Africa Albers Equal Area CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# Define WGS84 CRS for GPS coordinates
wgs84_crs <- st_crs(4326)

# Loop through each dataset in the list
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X", "Y"), crs = wgs84_crs)  # Convert to sf object with decimal degrees (WGS84) coordinate
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinates (in meters in Africa Albers)
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}
```

### Kenya 

####Laikipia

```{r}
# Laikipia
datasets <- c("LKlayout2025") 

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 1:3] # Keep only the first 3 columns LOC_ID, X, Y
  obj <- obj[-c(1), ] # Remove the first row - it is blank
  colnames(obj) <- as.character(obj[1, ]) # Set new row 1 as column names
  obj <- obj[-1, ] # Remove the new first row which repeats the column names
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c("LKlayout2025_spatial") 

# Project into Africa Albers
# Define Africa Albers Equal Area CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# UTM Zone 37N CRS
utm_37n_crs <- st_crs(32637)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("X_UTM37N", "Y_UTM37N"), crs = utm_37n_crs) # Convert to sf object using the UTM 37N coordinates
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinate
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}
```

#### Maasai Mara

```{r}
# Maasai Mara ~ format very different so going to import another dataset and use that
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/BHP")
MMspatial2018 <- read.csv("All.csv", header = TRUE)

datasets <- c("MMspatial2018") 

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  colnames(obj)[1] <- "LOC_ID"  # Rename first column to LOC_ID
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# List new datasets
datasets <- c("MMspatial2018_spatial") 

# Project into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")

# Decimal degrees CRS = WGS84
dd_crs <- st_crs(4326)

# Loop through each dataset to project to Africa Albers
for (ds in datasets) {
  df <- get(ds)
  df_sf <- st_as_sf(df, coords = c("Longitude", "Latitude"), crs = dd_crs)  # Convert to sf object using lon/lat in decimal degrees
  df_aea <- st_transform(df_sf, crs = africa_albers_crs) # Transform to Africa Albers
  coords_aea <- st_coordinates(df_aea) # Extract numeric coordinates
  df_final <- data.frame(LOC_ID = df$LOC_ID,
                         X_AfricaAlbers = coords_aea[, 1],
                         Y_AfricaAlbers = coords_aea[, 2]) # Combine LOC_ID and new coordinates
  assign(ds, df_final, envir = .GlobalEnv) # Overwrite the original dataset with the transformed coordinates
}
```

#### Ol Jogi

```{r}
datasets <- c("OJlayout2022")

# Loop through sites to extract spatial data 
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[, 3:5] # Keep only the first 3 columns LOC_ID, X, Y
  obj <- obj[, c(1, 3, 2)]           # Swap columns 2 and 3
  colnames(obj)[1] <- "LOC_ID" # Change column name
  new_name <- paste0(ds, "_spatial") # Save to a new object with "_spatial" suffix
  assign(new_name, obj, envir = .GlobalEnv)
}

# Remove duplicates
OJlayout2022_spatial <- OJlayout2022_spatial[!duplicated(OJlayout2022_spatial$LOC_ID), ]

# Define CRS
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +datum=WGS84 +units=m +no_defs")

# Convert, transform, and extract coords
df_sf <- st_as_sf(OJlayout2022_spatial, coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)
df_aea <- st_transform(df_sf, africa_albers_crs)
coords_aea <- st_coordinates(df_aea)

# Add numeric Africa Albers coords back to the original data.frame
OJlayout2022_spatial$X_AfricaAlbers <- coords_aea[,1]
OJlayout2022_spatial$Y_AfricaAlbers <- coords_aea[,2]

# Remove original lon/lat
OJlayout2022_spatial$Longitude <- NULL
OJlayout2022_spatial$Latitude  <- NULL
```

### Save spatial datasets to drive

```{r}
datasets <- c("RRlayoutMIO2018_spatial", "RRlayoutRNP2018_spatial", "RRlayoutWMA2018_spatial", "RRlayoutRGI2019_spatial", "SNlayoutNMTL2020_spatial", "SNlayoutSKPE2021_spatial", "SNlayoutSKPW2021_spatial", "SNlayoutSLWN2021_spatial", "SNlayoutSMGN2021_spatial", "SNlayoutNKLC2022_spatial", "SNlayoutNMSK2022_spatial", "ZBlayoutLL2020_spatial", "ZBlayoutLL2021_spatial", "ZBlayoutLL2022_spatial", "ZBlayoutR2020_spatial",  "ZBlayoutR2021_spatial", "ZBlayoutR2022_spatial", "LKlayout2025_spatial", "MMspatial2018_spatial", "OJlayout2022_spatial")

# Define output directory
output_dir <- "C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all datasets
for (ds in datasets) {
  file_name <- paste0(ds, ".csv")
  file_path <- file.path(output_dir, file_name)
  write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

## Prepare layout files for detection histories  

### Tanzania

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022")

# Loop through to remove location data and end day date
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[-c(2,3), -c(2, 3)] # Remove row 2 & 3 and columns 2 & 3
  obj[1, ] <- sub(" .*", "", obj[1, ]) # Keep only date part in row 1
  colnames(obj) <- as.character(obj[1, ]) # Set row 1 as column names
  obj <- obj[-1, ] # Remove the new first row
  assign(ds, obj, envir = .GlobalEnv) # Assign cleaned dataset back to the environment
}
```

### Zambia

```{r}
datasets <- c("ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022")

# Append year onto loc-IDs due to location inconsistencies
ZBlayoutLL2020 <- ZBlayoutLL2020 %>%
  mutate(Grid.cell = paste0(Grid.cell, "-2020"))
ZBlayoutLL2021 <- ZBlayoutLL2021 %>%
  mutate(Grid.cell = paste0(Grid.cell, "-2021"))
ZBlayoutLL2022 <- ZBlayoutLL2022 %>%
  mutate(Grid.cell = paste0(Grid.cell, "-2022"))
ZBlayoutR2020 <- ZBlayoutR2020 %>%
  mutate(Grid.cell = paste0(Grid.cell, "-2020"))
ZBlayoutR2021 <- ZBlayoutR2021 %>%
  mutate(Si1e.ID = paste0(Si1e.ID, "-2021"))
ZBlayoutR2022 <- ZBlayoutR2022 %>%
  mutate(Grid.Cell.ID = paste0(Grid.Cell.ID, "-2022"))

# Loop through and modify each dataset to remove location data and match date format to Tanzania: DD.MM.YYYY -> DD/MM/YYYY and remove name of first column LOC_ID
for (ds in datasets) {
  df <- get(ds)
  df <- df[-c(2,3), -c(2, 3, 4)] # Remove row 2 & 3 and columns 2 & 3
  new_names <- colnames(df) 
  date_cols <- grepl("^\\d{2}\\.\\d{2}\\.\\d{4}$" , new_names)# Identify which column names match the date pattern
  new_names[date_cols] <- gsub("\\.", "/", new_names[date_cols]) # Replace dots with slashes in those column names
  colnames(df) <- new_names  # Apply updated names and overwrite original dataset
  names(df)[1] <- ""  # Remove name of first column
  assign(ds, df, envir = .GlobalEnv)
}


#ZBlayoutLL2020 <- ZBlayoutLL2020[, -98]
```

### Kenya 

#### Laikipia

```{r}
datasets <- c("LKlayout2025")

# Loop through to remove location data and end day date
for (ds in datasets) {
  obj <- get(ds)
  obj <- obj[-c(1,2), -c(2, 3, 4, 5)]  # Remove row 1 & 2 and columns 2,3,4,5
  colnames(obj)[1] <- ""  # Remove the column name of the first column
  colnames(obj) <- gsub("\\.", "/", sub("^X(\\d{2})\\.(\\d{2})\\.(\\d{4}).*", "\\1.\\2.\\3", colnames(obj)))  # Convert XDD.MM.YYYY.* to DD/MM/YYYY
  assign(ds, obj, envir = .GlobalEnv)  # Assign cleaned dataset back to the environment
}
```

#### Maasai Mara

```{r}
MMlayout2018 <- as.data.frame(t(MMlayout2018)) 
colnames(MMlayout2018) <- MMlayout2018[1, ]  # set first row as column names
MMlayout2018 <- MMlayout2018[-1, ]           # remove that first row
colnames(MMlayout2018) <- format(as.Date(colnames(MMlayout2018), format = "%Y.%m.%d"),"%d/%m/%Y") # change date format
MMlayout2018$RowName <- rownames(MMlayout2018)  # copy row names into a new column
rownames(MMlayout2018) <- NULL                  # remove row names
MMlayout2018 <- MMlayout2018[, c(ncol(MMlayout2018), 1:(ncol(MMlayout2018)-1))]  # move RowName to column 1
colnames(MMlayout2018)[1] <- "" # remove column name
MMlayout2018[is.na(MMlayout2018)] <- 0 # change all NAs to 0
MMlayout2018[] <- lapply(MMlayout2018, function(x) {if (is.character(x)) sub("^\\s+1$", "1", x) else x}) # clean spaces
```

#### Ol Jogi - TO ADD

```{r}
datasets <- c("")


```

### Save clean layout datasets

```{r}
datasets <- c("RRlayoutMIO2018", "RRlayoutRNP2018", "RRlayoutWMA2018", "RRlayoutRGI2019", "SNlayoutNMTL2020", "SNlayoutSKPE2021", "SNlayoutSKPW2021", "SNlayoutSLWN2021", "SNlayoutSMGN2021", "SNlayoutNKLC2022", "SNlayoutNMSK2022","ZBlayoutLL2020", "ZBlayoutLL2021", "ZBlayoutLL2022", "ZBlayoutR2020",  "ZBlayoutR2021", "ZBlayoutR2022", "LKlayout2025", "MMlayout2018")

# Define output directory
output_dir <- "C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all objects in the environment
for (ds in datasets) {
    file_name <- paste0(ds, "_clean.csv")
    file_path <- file.path(output_dir, file_name)
    write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

# Species pre-processing

## Import raw detection data files
```{r}
# read Ruaha-Rungwa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Ruaha-Rungwa_CTcapture_data")

RRdataMIO2018 <- read.csv("TZ_RR_MIO-2018_CT captures_202408.csv", header = FALSE)
RRdataRNP2018 <- read.csv("TZ_RR_RNP-2018_CT captures_202408.csv", header = FALSE)
RRdataWMA2018 <- read.csv("TZ_RR_WMA-2018_CT captures_202408.csv", header = FALSE)
RRdataRGI2019 <- read.csv("TZ_RR_RGI-2019_CT captures_202408.csv", header = FALSE)

# read Selous-Nyerere data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Selous-Nyere_CTcapture_data")

SNdataNMTL2020 <- read.csv("TZ_SN_NMTL-2020_CT captures_202405.csv", header = FALSE)
SNdataSKPE2021 <- read.csv("TZ_SN_SKPE-2021_CT captures_202405.csv", header = FALSE)
SNdataSKPW2021 <- read.csv("TZ_SN_SKPW-2021_CT captures_202405.csv", header = FALSE)
SNdataSLWN2021 <- read.csv("TZ_SN_SLWN-2021_CT captures_202405.csv", header = FALSE)
SNdataSMGN2021 <- read.csv("TZ_SN_SMGN-2021_CT captures_202405.csv", header = FALSE)
SNdataNKLC2022 <- read.csv("TZ_SN_NKLC-2022_CT captures_202405.csv", header = FALSE)
SNdataNMSK2022 <- read.csv("TZ_SN_NMSK-2022_CT captures_202405.csv", header = FALSE)

# read Lower Luangwa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Lower_Luangwa_CTcapture_data")

ZBdataLL2020 <- read.csv("All_Species_2020.csv", header = TRUE)
ZBdataLL2021 <- read.csv("All_Species_2021.csv", header = TRUE)
ZBdataLL2022 <- read.csv("LLZ_BM_CT_LowerLuangwa_2022.csv", header = TRUE)

# read Rufunsa data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Rufunsa_CTcapture_data")

ZBdataR2020 <- read.csv("All_Species_2020.csv", header = TRUE)
ZBdataR2021 <- read.csv("All_Species_2021.csv", header = TRUE)
ZBdataR2022 <- read.csv("All_Species_2022.csv", header = TRUE)

# read Laikipia data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Laikipia")

LKdata2025 <- read.csv("Camera_trap_uncleaned2.csv", header = TRUE)

# read Maasai Mara data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/BHP")

MMdata2018 <- read.csv("ML2018.csv", header = TRUE)
MMdatahyaena <- read.csv("hyena_aardwolf_2018_ML.csv", header = TRUE)
MMhyaena <- read.csv("Hyena-export.csv", header = TRUE)

# read Ol Jogi data
setwd("~/3. Thesis/Chapter 3 - Co-occcurence/Data/Ol Jogi")
OJdata2022 <- read.csv("Oljogi.csv", header = TRUE)
```

## Prepare detection data as lists and count detections per site x day x species

### Tanzania

```{r}
# List of datasets
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022")

# Loop over the datasets and set second row as column names and convert DATE to Date class
for (ds in datasets) {
  df <- get(ds) # Access the dataset using get() to modify it
  df <- df[-1, ] # Remove the first row
  colnames(df) <- df[1, ] # Set the second row as column names
  df <- df[-1, ] # Remove the second row (now the header)
  df$DATE <- as.Date(df$DATE, format = "%d/%m/%Y") # Convert DATE column to Date class
  assign(ds, df) # Assign the cleaned data back to the original dataset name
}

# Count detections per species per day per site
species <- c("lion", "cheetah", "leopard", "spotted hyaena", "striped hyaena", 
             "wild dog", "caracal", "serval", "civet", "honey badger", 
             "wildcat", "bat-eared fox", "aardwolf", "b-b jackal", "s-s jackal")
             
# Initialize an empty list to store the results
result_countsTZ <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    df <- get(ds) # get the dataset and use [[ to access the column
    filtered_data <- df[df[[sp]] == 1, ] # Filter the dataset for rows where the species column has a value of 1
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Get the unique combinations of DATE and LOC_ID
    total_unique_combinations <- nrow(unique_combinations) # Count the total number of unique combinations
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create a result name: dataset name_species
    result_countsTZ[[result_name]] <- total_unique_combinations # Store the result in the list with the result name
  }
}

# Extract 1 species to double check outputs (RRMIO2018 Lion - 46)
lion_data <- RRdataMIO2018[RRdataMIO2018$lion == 1, ] # Filter the dataset for rows where the lion column has a value of 1
unique_combinations <- unique(lion_data[, c("DATE", "LOC_ID")]) # Get the unique combinations of DATE and LOC_ID
total_unique_combinations <- nrow(unique_combinations) # Count the total number of unique combinations
total_unique_combinations # Print the result

# Turn into list of detections rather than a matrix

# Loop through and subset each dataset to remove unwanted species and pivot into one row per detection
for (ds in datasets) {
  df <- get(ds)
  cols_to_keep <- c(1:8, which(colnames(df) %in% species)) # Get all matching columns: first 8 + any from species list
  cols_to_keep <- unique(cols_to_keep)  # Avoid duplicates
  df <- df[, cols_to_keep]
  species_cols <- intersect(colnames(df), species) # Identify columns that are in the species list
  df <- df[!apply(df[, species_cols], 1, function(row) all(as.numeric(row) == 0)), ] # Remove rows where all remaining species columns are 0
    df <- df %>%
    pivot_longer(
      cols = all_of(species_cols),
      names_to = "Species",
      values_to = "value"
    ) %>%
    filter(as.numeric(value) == 1) # Pivot longer and filter for presence == 1
  assign(ds, df, envir = .GlobalEnv)
}
```
 
### Zambia

#### Lower Luangwa  

```{r}
# NOTE: These are processed separately as there are different issues for each file

# LL2020 cleaning
ZBdataLL2020 <- ZBdataLL2020 %>% rename(Species = Tags, DATE = Date, LOC_ID = Area) # Rename columns To = From
ZBdataLL2020 <- ZBdataLL2020 %>% mutate(Species = recode(Species,
    "African Wild Dog" = "Wild Dog - Lycaon pictus",
    "Wild Dog - Canis pictus" = "Wild Dog - Lycaon pictus",
    "Spotted Hyena - Crocuta crocuta" = "Spotted Hyaena - Crocuta crocuta"
  )) # Rename species
ZBdataLL2020$LOC_ID <- gsub("[A-Za-z]", "", ZBdataLL2020$LOC_ID) # Clean LOC_ID to match layout dataframe
ZBdataLL2020$DATE <- as.Date(ZBdataLL2020$DATE, format = "%d/%m/%Y") # Match date formats

# LL2021 cleaning
ZBdataLL2021 <- ZBdataLL2021 %>% rename(DATE = Date.and.Time, LOC_ID = Cell) # Rename columns
ZBdataLL2021 <- ZBdataLL2021 %>% # Pivot for multiple species columns into 1
  pivot_longer(
    cols = c(Species.1, Species.2, Species.3),
    names_to = "Species_Column",
    values_to = "Species"
  ) %>%
  filter(!is.na(Species) & Species != "" & Species != "NA")  # Remove rows where Species was NA
ZBdataLL2021$Species[ZBdataLL2021$Species == "Honeybadger - Mellivora capensis"] <-"Honey Badger - Mellivora capensis" # Rename species
ZBdataLL2021$DATE <- as.Date(ZBdataLL2021$DATE, format = "%d/%m/%Y") # Match date formats

# LL2022 cleaning
# Split the data into Photo, Date, Time, Tag
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(
    Photo = str_extract(PhotoDateTimeTag, "^.*(?=\\d{2}/\\d{2}/\\d{2})"), # Extract Photo (everything before the date
    DATE = str_extract(PhotoDateTimeTag, "\\d{2}/\\d{2}/\\d{2}"), # Extract Date (DD/MM/YY)
    Time = str_extract(PhotoDateTimeTag, "\\d{2}:\\d{2}:\\d{1,2}(?: [APM]{2})?"),# Extract Time (handles formats like HH:MM:S or HH:MM:SS, optional AM/PM)
    Species = str_extract(PhotoDateTimeTag, "(?<=\\d{2}:\\d{2}:\\d{1,2}(?: [APM]{2})?).*") %>% 
      str_trim() # Extract Species
  )
  
# Clean species names in LL2022 ~ do this pre splitting as it makes splitting easier
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = str_replace_all(Species, "\\d+_|LHS_|LS_|RS_|RHS_|_RHS|_LHS|Front\\s+end_|Back\\s+end_|Highlight_|Hghlight_|_Sub-adult|_Highlight|_White|White_|IR_|_IR|_Snare|Juvenile_|_Juvenile|2 indiv_|_2 indiv|_Wounded|Adult_|\\b\\d{1,2} [AP]M(?=\\w)|^_|\\s{2,}", ""), # Remove specific substrings from anywhere in the string | times eg."2 PM" or "11 AM" even when directly followed by a word | leading underscore | double spaces
  )

# Clean specific names pre-splitting
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = case_when(
    Species %in% c("African - Civettictis civetta_Civet", "Civet African - Civettictis civetta", "ivet African - Civettictis civetta", "Cvet African - Civettictis civetta") ~ "African Civet - Civettictis civetta",
    Species %in% c("Jackal_Side-Striped - Canis adustus", "Jackal Side-Striped - Canis adustus") ~ "Side-Striped Jackal - Canis adustus",
    Species %in% c("Hyaena_Spotted  - Crocuta crocuta", "Hyaena Spotted  - Crocuta crocuta", "Hyaena_Spotted - Crocuta crocuta") ~ "Spotted Hyaena - Crocuta crocuta",
    Species == "Honeybadger - Mellivora capensis" ~ "Honey Badger - Mellivora capensis",
    Species == "Wild cat African - Felis lybica" ~ "African Wildcat - Felis lybica",
    Species %in% c("dult_Wild Dog - Lycaon pictus", "uvenile_Wild Dog - Lycaon pictus", "ild Dog - Lycaon pictus", "Wld Dog - Lycaon pictus", "Back end_Wild Dog - Lycaon pictus") ~ "Wild Dog - Lycaon pictus",
    TRUE ~ Species
  ))

# Split the Species column for multiple species and pivot
ZBdataLL2022 <- ZBdataLL2022 %>%
  separate_wider_delim(Species, delim = "_", names_sep = "_part", too_few = "align_start") %>%
  pivot_longer(
    cols = c(Species_part1, Species_part2, Species_part3, Species_part4),
    names_to = "Species_Column",
    values_to = "Species"
  ) %>%
  filter(!is.na(Species) & Species != "" & Species != "NA")  # Remove rows where Species was NA

# Extra cleaning post splitting
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(Species = case_when(
    Species %in% c("Hyaena Spotted- Crocuta crocuta", "Spotted- Crocuta crocuta") ~ "Spotted Hyaena - Crocuta crocuta",
    Species == "Civet African - Civettictis civetta" ~ "African Civet - Civettictis civetta",
    TRUE ~ Species
  ))

# Loc ID extraction - needs further cleaning
ZBdataLL2022 <- ZBdataLL2022 %>% mutate(
    LOC_ID = str_extract(Photo, "^.*?(?=R)"),              # Extract everything before 'R'
    LOC_ID = gsub("[A-Za-z_]", "", LOC_ID)                 # Remove letters and underscores
  )

# Match date formats
ZBdataLL2022$DATE <- as.Date(ZBdataLL2022$DATE, format = "%d/%m/%y")


                       # Count number of days with detections of each species per site

datasets <- c("ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022")
species <- c("Lion - Panthera leo", "cheetah", "Leopard - Panthera pardus", "Spotted Hyaena - Crocuta crocuta", "Wild Dog - Lycaon pictus", "Caracal - Caracal caracal", "Serval - Leptailurus serval", "African Civet - Civettictis civetta", "Honey Badger - Mellivora capensis", "African Wildcat - Felis lybica",  "Side-Striped Jackal - Canis adustus")

# Initialize an empty list to store the results
result_countsLL <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp) # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result name
    result_countsLL[[result_name]] <- total_unique_combinations # Store result
  }
}

# Append year onto loc-IDs due to location inconsistencies
ZBdataLL2020 <- ZBdataLL2020 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2020"))
ZBdataLL2021 <- ZBdataLL2021 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2021"))
ZBdataLL2022 <- ZBdataLL2022 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2022"))
```

#### Rufunsa

```{r}
# R2020 cleaning
ZBdataR2020 <- ZBdataR2020 %>% rename(Species = Species.1) # Rename column - no other species columns
ZBdataR2020$Species[ZBdataR2020$Species == "Spotted Hyena - Crocuta crocuta"] <-"Spotted Hyaena - Crocuta crocuta" #Rename Species
ZBdataR2020$DATE <- as.Date(ZBdataR2020$DATE, format = "%d/%m/%Y") # Match date format

# R2021 cleaning
ZBdataR2021 <- ZBdataR2021 %>% rename(Species = Species.1) # Other species column has no carnivores
ZBdataR2021$Species[ZBdataR2021$Species == "Honeybadger - Mellivora capensis"] <-"Honey Badger - Mellivora capensis" #Rename species
ZBdataR2021 <- ZBdataR2021 %>%
  mutate(
    DATETIME = as.POSIXct(DATETIME, format = "%d/%m/%Y", tz = "UTC"), 
    DATE = as.Date(DATETIME) # Extract and create DATE column
  )

# R2022 cleaning
ZBdataR2022 <- ZBdataR2022 %>% rename(Species = Species.1, Checkdate = DATE) # Rename columns
ZBdataR2022 <- ZBdataR2022 %>%
  mutate(
    DATETIME = as.POSIXct(DATETIME, format = "%d/%m/%Y", tz = "UTC"),  
    DATE = as.Date(DATETIME) # Extract and create DATE column
  )
ZBdataR2022 <- ZBdataR2022 %>%
  mutate(Species = recode(Species,
    "Honeybadger - Mellivora capensis" = "Honey Badger - Mellivora capensis",
    "Civet African - Civettictis civetta" = "African Civet - Civettictis civetta",
    "Jackal Side-Striped - Canis adustus" = "Side-Striped Jackal - Canis adustus",
    "Hyaena Spotted  - Crocuta crocuta" = "Spotted Hyaena - Crocuta crocuta"
  ))

# Count number of days with detections of each species per site
datasets <- c("ZBdataR2020", "ZBdataR2021", "ZBdataR2022")
species <- c("Lion - Panthera leo", "Leopard - Panthera pardus", "Spotted Hyaena - Crocuta crocuta", "Wild Dog - Lycaon pictus", "Caracal - Caracal caracal", "Serval - Leptailurus serval", "African Civet - Civettictis civetta", "Honey Badger - Mellivora capensis", "African Wildcat - Felis lybica", "Side-Striped Jackal - Canis adustus")

# Initialize an empty list to store the results
result_countsR <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp)  # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result nam
    result_countsR[[result_name]] <- total_unique_combinations # Store result
  }
}

# Add year to location names due to inconsistencies
ZBdataR2020 <- ZBdataR2020 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2020"))
ZBdataR2021 <- ZBdataR2021 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2021"))
ZBdataR2022 <- ZBdataR2022 %>%
  mutate(LOC_ID = paste0(LOC_ID, "-2022"))
```

### Kenya 

#### Laikipia

```{r}
names(LKdata2025)[names(LKdata2025) == "Image.ID"] <- "PHOTO"  # Rename Image.ID to PHOTO

LKdata2025 <- cbind(LKdata2025[1], GRID = "LKE", LKdata2025[-1])  # Insert new column 2 called GRID with value LKE

names(LKdata2025)[names(LKdata2025) == "Site.Name"] <- "LOC_ID"  # Rename Site.Name to LOC_ID

ts_col <- if ("Timestamp" %in% names(LKdata2025)) "Timestamp" else if ("Timestamp" %in% names(LKdata2025)) "Timestamp" else NA  # Find timestamp column

ts_vals <- gsub("T", " ", as.character(LKdata2025[[ts_col]]))  # Replace T with space for splitting
raw_date <- sub(" .*", "", ts_vals)  # Extract DATE part (DD/MM/YYYY)
LKdata2025$DATE <- format(as.Date(raw_date, format = "%d/%m/%Y"), "%Y-%m-%d")  # Convert to YYYY-MM-DD

LKdata2025$TIME <- sub("^[^ ]* ?", "", ts_vals)  # Extract TIME from timestamp

names(LKdata2025)[names(LKdata2025) == "Timestamp"] <- "DATETIME"  # Rename timestamp to DATETIME

LKdata2025$value <- 1  # Add new column 'value' at the end with all values = 1

LKdata2025 <- LKdata2025[ , !names(LKdata2025) %in% "Camera.Name" ]  # Remove column Camera.Name

LKdata2025 <- LKdata2025 %>% # split rows with two species
  separate_rows(Species, sep = "\\|") %>%
  mutate(Species = str_trim(Species))

# Count number of days with detections of each species per site
datasets <- c("LKdata2025")
species <- c("Aardwolf", "Bat-eared Fox", "Black-backed jackal", "Caracal", "Cheetah", "Honey badger", "Leopard", "Lion", "Serval", "Side-striped jackal", "Spotted hyena", "Striped hyena", "Wild Dog", "Wildcat")

# Initialize an empty list to store the results
result_countsLK <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp)  # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result nam
    result_countsLK[[result_name]] <- total_unique_combinations # Store result
  }
}
```

#### Maasai Mara

```{r}
# Remove species not needed and set up main dataset to match
keep_species <- c(
  "serval",
  "jackal",
  "lion",
  "bateared_fox",
  "cheetah",
  "leopard"
)

MMdata2018 <- MMdata2018[MMdata2018$pred1_label %in% keep_species, ]

MMdata2018 <- MMdata2018[, c("img_id", "pred1_label", "location", "datetime", "date")] # remove unwanted columns

colnames(MMdata2018) <- c("PHOTO", "Species", "LOC_ID", "DATETIME", "DATE") # rename
MMdata2018 <- MMdata2018[, c("PHOTO", "LOC_ID", "Species", "DATETIME", "DATE")] # reorder

MMdata2018 <- cbind(MMdata2018[1], GRID = "MM", MMdata2018[-1])  # Insert new column 2 called GRID with value MM
MMdata2018$value <- 1 # add value column for count
MMdata2018$TIME <- sub(".* ", "", MMdata2018$DATETIME)  # extract time after the space
MMdata2018 <- MMdata2018[, c("PHOTO", "GRID", "LOC_ID", "Species", "DATETIME", "DATE", "TIME", "value")] # reorder

# edit hyaena data 
keep_species <- c("hyena_spotted", "aardwolf", "hyena_striped") # remove unwanted species
MMhyaena <- MMhyaena[MMhyaena$label %in% keep_species, ]

# merge hyaena split to hyaena metadata and set up to match main data
MM_joined <- inner_join(MMdatahyaena, MMhyaena, by = "img_id") # join metadata

MM_joined <- MM_joined[, c("img_id", "label", "location", "datetime", "date")] 
colnames(MM_joined) <- c("PHOTO", "Species", "LOC_ID", "DATETIME", "DATE") # rename

MM_joined$DATETIME <- format(as.POSIXct(MM_joined$DATETIME, format = "%d/%m/%Y %H:%M"), "%Y-%m-%d %H:%M:%S") # reformat DATETIME
MM_joined$DATE <- format(as.Date(MM_joined$DATE, format = "%d/%m/%Y"), "%Y-%m-%d") # reformat DATE

MM_joined$GRID <- "MM" # Insert new column called GRID with value MM
MM_joined$value <- 1  # add value column for count
MM_joined$TIME <- sub(".* ", "", MM_joined$DATETIME)  # extract time after the space
MM_joined <- MM_joined[, c("PHOTO", "GRID", "LOC_ID", "Species", "DATETIME", "DATE", "TIME", "value")] # reorder

# join to main
MMdata2018 <- bind_rows(MMdata2018, MM_joined)

# fix jackal now to avoid later complications
MMdata2018$Species[MMdata2018$Species == "jackal"] <- "Black-backed jackal"

# Count number of days with detections of each species per site
datasets <- c("MMdata2018")
species <- c("aardwolf", "bateared_fox", "Black-backed jackal", "cheetah", "hyena_spotted", "hyena_striped", "leopard", "lion", "serval")

# Initialize an empty list to store the results
result_countsMM <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp)  # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result nam
    result_countsMM[[result_name]] <- total_unique_combinations # Store result
  }
}
```

#### Ol Jogi

```{r}
# Remove unnecessary columns
OJdata2022 <- OJdata2022[, c("Image.ID", "Site.Name", "Latitude", "Longitude", "Species", "Timestamp")] 

# Rename current columns to match
colnames(OJdata2022) <- c("PHOTO", "LOC_ID", "Lat", "Long", "Species", "DATETIME")

# Create date and time columns
OJdata2022$DATETIME <- as.POSIXct(OJdata2022$DATETIME, format = "%d/%m/%Y %H:%M")
OJdata2022$DATE <- as.Date(OJdata2022$DATETIME)
OJdata2022$TIME <- format(OJdata2022$DATETIME,  "%H:%M:%S")
OJdata2022$GRID <- "OJ"
OJdata2022$value <- 1
OJdata2022 <- OJdata2022[, c("PHOTO", "GRID", "LOC_ID", "Species", "DATETIME", "DATE", "TIME", "value")]

# Count number of days with detections of each species per site
datasets <- c("OJdata2022")
species <- c("Aardwolf", "Bat-eared Fox", "Black-backed jackal", "Caracal", "Leopard", "Lion", "Ratel", "Serval", "Spotted hyeana", "Striped Hyaena", "Wild Dog", "Wildcat")

# Initialize an empty list to store the results
result_countsOJ <- list()

# Loop over species and datasets
for (sp in species) {
  for (ds in datasets) {
    dataset <- get(ds)
    filtered_data <- subset(dataset, Species == sp)  # Filter rows where Species matches current species
    unique_combinations <- unique(filtered_data[, c("DATE", "LOC_ID")]) # Count unique DATE × LOC_ID combinations
    total_unique_combinations <- nrow(unique_combinations)
    result_name <- paste0(ds, "_", gsub(" ", "_", sp)) # Create result nam
    result_countsOJ[[result_name]] <- total_unique_combinations # Store result
  }
}
```


### Match species names across datasets and remove unneccsary species data

```{r}
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022", "ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022", "ZBdataR2020", "ZBdataR2021", "ZBdataR2022", "LKdata2025", "MMdata2018", "OJdata2022" )

# Create a named vector for recoding
species_recode <- c(
  "Lion - Panthera leo" = "Lion",
  "lion" = "Lion",
  "Leopard - Panthera pardus" = "Leopard",
  "leopard" = "Leopard",
  "cheetah" = "Cheetah",
  "Spotted Hyaena - Crocuta crocuta" = "Spotted hyaena",
  "hyena_spotted" = "Spotted hyaena",
  "spotted hyaena" = "Spotted hyaena",
  "Spotted hyena" = "Spotted hyaena",
  "Spotted hyeana" = "Spotted hyaena",
  "striped hyaena" = "Striped hyaena",
  "Striped hyena" = "Striped hyaena",
  "hyena_striped" = "Striped hyaena",
  "Striped Hyaena" = "Striped hyaena",
  "Wild Dog - Lycaon pictus" = "African wild dog",
  "Wild Dog" = "African wild dog",
  "wild dog" = "African wild dog",
  "Caracal - Caracal caracal" = "Caracal",
  "caracal" = "Caracal",
  "Serval - Leptailurus serval" = "Serval",
  "serval" = "Serval",
  "African Civet - Civettictis civetta" = "African civet",
  "civet" = "African civet",
  "Honey Badger - Mellivora capensis" = "Honey badger",
  "Ratel" = "Honey badger",
  "honey badger" = "Honey badger",
  "African Wildcat - Felis lybica" = "African wildcat",
  "wildcat" = "African wildcat",
  "Wildcat" = "African wildcat",
  "Side-Striped Jackal - Canis adustus" = "Side-striped jackal",
  "s-s jackal" = "Side-striped jackal",
  "bat-eared fox" = "Bat-eared fox",
  "bateared_fox" = "Bat-eared fox",
  "Bat-eared Fox" = "Bat-eared fox",
  "Bat-eared Fox" = "Bat-eared fox",
  "aardwolf" = "Aardwolf",
  "b-b jackal" = "Black-backed jackal"
)

# Loop through all datasets and apply recoding
for (ds in datasets) {
  df <- get(ds)
  df <- df %>%
    mutate(Species = recode(Species, !!!species_recode)) %>% # Recode species
    filter(Species %in% unname(species_recode))  # Keep only recoded species
  assign(ds, df, envir = .GlobalEnv)
}
```

### Save detection datasets

```{r}
# Save clean datasets
datasets <- c("RRdataMIO2018", "RRdataRNP2018", "RRdataWMA2018", "RRdataRGI2019", "SNdataNMTL2020", "SNdataSKPE2021", "SNdataSKPW2021", "SNdataSLWN2021", "SNdataSMGN2021", "SNdataNKLC2022", "SNdataNMSK2022", "ZBdataLL2020", "ZBdataLL2021", "ZBdataLL2022", "ZBdataR2020", "ZBdataR2021", "ZBdataR2022","LKdata2025", "MMdata2018", "OJdata2022")

# Define output directory
output_dir <- "C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis"

# Loop through all objects in the environment
for (ds in datasets) {
    file_name <- paste0(ds, "_clean.csv")
    file_path <- file.path(output_dir, file_name)
    write.csv(get(ds), file = file_path, row.names = FALSE)
}
```

# Spatial covariate pre-processing

## Import camera trap locations and combine and create buffer

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis")
datasets <- list.files(pattern = "\\_spatial.csv$")
sp_list <- lapply(datasets, read.csv)  
camlocs <- do.call(rbind, sp_list) # Merge 
camlocs_vec <- vect(camlocs, geom=c("X_AfricaAlbers", "Y_AfricaAlbers"), crs="+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25") # Convert to spatial file

# View data to sense check
mapview(camlocs_vec)

# Create a buffer of 1000 m around the camera trap locations
cam_1000m_buffer <- buffer(camlocs_vec, width = 1000)
mapview(cam_1000m_buffer)

# Create a buffer of 10,000 m around the camera trap locations
cam_10km_buffer <- buffer(camlocs_vec, width = 10000)
mapview(cam_10km_buffer)
```

## Temperature

### Prepare raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Temperature/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("temp"), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting temperature rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
temp_aa <- project(temp, africa_albers_crs, method = "near")

# Check they match
plot(temp_aa)
plot(study_vect, add=T)

# Mask
temp_mask <- mask(temp_aa, study_vect)
temp_crop <- crop(temp_mask, study_vect)
plot(temp_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(temp_crop, "temp_mask.tiff", overwrite=T)
```

## Temp seasonality

### Prepare raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Temp_seasonality/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("temp_seas"), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
temp_seas_aa <- project(temp_seas, africa_albers_crs, method = "near")

# Check they match
plot(temp_seas_aa)
plot(study_vect, add=T)

# Mask & crop
temp_seas_mask <- mask(temp_seas_aa, study_vect)
temp_seas_crop <- crop(temp_seas_mask, study_vect)
plot(temp_seas_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(temp_seas_crop, "temp_seas_mask.tiff", overwrite=T)
```

## Land temp

### Prepare raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Land temp/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("land_temp", i), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
land_temp_aa1 <- project(land_temp1, africa_albers_crs, method = "near")
land_temp_aa2 <- project(land_temp2, africa_albers_crs, method = "near")

# Merging landcover rasters
land_temp_aa <- merge(land_temp_aa1, land_temp_aa2)

# Check they match
plot(land_temp_aa)
plot(study_vect, add=T)

# Mask & crop
land_temp_mask <- mask(land_temp_aa, study_vect)
land_temp_crop <- crop(land_temp_mask, study_vect)
plot(land_temp_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(land_temp_crop, "land_temp_mask.tiff", overwrite=T)
```

## Rainfall

### Prepare raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Precipitation/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("precip"), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
precip_aa <- project(precip, africa_albers_crs, method = "near")

# Check they match
plot(precip_aa)
plot(study_vect, add=T)

# Mask & crop
precip_mask <- mask(precip_aa, study_vect)
precip_crop      <- crop(precip_mask, study_vect)
plot(precip_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(precip_crop, "precip_mask.tiff", overwrite=T)
```

## Rain seasonality

### Prepare raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Precip_seasonality/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("precip_seas"), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
precip_seas_aa <- project(precip_seas, africa_albers_crs, method = "near")

# Check they match
plot(precip_seas_aa)
plot(study_vect, add=T)

# Mask & crop
precip_seas_mask <- mask(precip_seas_aa, study_vect)
precip_seas_crop <- crop(precip_seas_mask, study_vect)
plot(precip_seas_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(precip_seas_crop, "precip__seas_mask.tiff", overwrite=T)
```

## Distance to water

### Prepare raster

```{r}
# read in land cover
lndc_mask <- rast("C:/R Projects/Co-occurence/lndc_mask.tiff") 

# make just water raster
water <- mask(lndc_mask, lndc_mask, maskvalues = 1, inverse = TRUE)
writeRaster(water, "water.tiff", overwrite=T)

# reproject to 50m resolution
water_50m <- aggregate(water, fact = 5, fun = max, na.rm = TRUE)

# calcualte distance and mask to study area
water_dist <- distance(water_50m, unit = "km")
water_dist <- mask(water_dist, study_vect)

writeRaster(water_dist, "water_dist.tiff", overwrite=T)
```

## Area protected classification

### Prepare raster

```{r}
# read in raster
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/WDPA/")
WDPA <- rast("WDPA_PolygonToRaster.tif")

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
WDPA_aa <- project(WDPA, africa_albers_crs, method = "near")

# Check they match
plot(WDPA_aa)
plot(study_vect, add=T)

# Mask & crop
WDPA_mask <- mask(WDPA_aa, study_vect)
WDPA_mask <- crop(WDPA_mask, study_vect)
plot(WDPA_mask)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

# Reclassify values
reclass <- matrix(
  c(0, NA,
    1, 6,
    2, NA,
    3, 4,
    4, 2,
    5, NA,
    6, 1,
    7, NA,
    8, 1,
    9, 3
  ),
  ncol = 2,
  byrow = TRUE
)

WDPA_mask <- classify(WDPA_mask, reclass)
plot(WDPA_mask)

writeRaster(WDPA_mask, "WDPA_mask.tiff", overwrite=T)
```

# Extract exact point covariates ~ skip to here if reimporting rasters

```{r}
# Read rasters back in
lndc_crop<- rast("C:/R Projects/Co-occurence/lndc_mask.tiff") 
temp_crop <- rast("C:/R Projects/Co-occurence/temp_mask.tiff") 
temp_seas_crop <- rast("C:/R Projects/Co-occurence/temp_seas_mask.tiff") 
precip_crop <- rast("C:/R Projects/Co-occurence/precip_mask.tiff") 
precip_seas_crop <- rast("C:/R Projects/Co-occurence/precip__seas_mask.tiff") 
land_temp_crop <-  rast("C:/R Projects/Co-occurence/land_temp_mask.tiff")
water_dist <- rast("C:/R Projects/Co-occurence/water_dist.tiff") 
WDPA_mask <- rast("C:/R Projects/Co-occurence/WDPA_mask.tiff") 

# create a new lndc and land_temp has exactly the same geometry as temp_crop
lndc_matched <- resample(lndc_crop, temp_crop, method = "near")
land_temp_matched <- resample(land_temp_crop, temp_crop, method = "near")
water_matched <- resample(water_dist, temp_crop, method = "near")
WDPA_matched <- resample(WDPA_mask, temp_crop, method = "near")

# Stack rasters
stack <- c(lndc_matched, temp_crop, temp_seas_crop, land_temp_matched, precip_crop, precip_seas_crop, water_matched, WDPA_matched) # add PA classification

# Extract to camera stations
values <- extract(stack, camlocs_vec, df=T, bind=T)
head(values)
names(values)[2] <- "land_cover"
names(values)[3] <- "temp"
names(values)[4] <- "temp_seas"
names(values)[5] <- "land_temp"
names(values)[6] <- "precip"
names(values)[7] <- "precip_seas"
names(values)[8] <- "water_dist"
names(values)[9] <- "WDPA"
head(values)
```

## Landcover

### Prepare raster - not needed if reimporting saved masked raster

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Landcover/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("lndc", i), rast(files[i]))
}

# Crop and Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting landcover rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
lndc1_aa <- project(lndc1, africa_albers_crs, method = "near")
lndc2_aa <- project(lndc2, africa_albers_crs, method = "near")
lndc3_aa <- project(lndc3, africa_albers_crs, method = "near")
lndc4_aa <- project(lndc4, africa_albers_crs, method = "near")
lndc5_aa <- project(lndc5, africa_albers_crs, method = "near")
lndc6_aa <- project(lndc6, africa_albers_crs, method = "near")
lndc7_aa <- project(lndc7, africa_albers_crs, method = "near")

# Merging landcover rasters
lndc_aa <- merge(lndc1_aa, lndc2_aa, lndc3_aa, lndc4_aa, lndc5_aa, lndc6_aa, lndc7_aa)

# Check they match
plot(lndc_aa)
plot(study_vect, add=T)

# Mask & crop & extend to match extents
lndc_mask <- mask(lndc_aa, study_vect)
lndc_crop <- crop(lndc_mask, study_vect)
plot(lndc_crop)
plot(study_vect,add=T)
plot(camlocs_vec, add=T)

writeRaster(lndc_crop, "lndc_mask.tiff", overwrite=T)
```

### Caclulate proportions within buffer

```{r}
# Read raster back in
lndc_mask <- rast("C:/R Projects/Co-occurence/lndc_mask.tiff") 

# Extract raster values inside each buffer polygon
lndc_vals <- extract(lndc_mask, cam_1000m_buffer)

# Create a lookup table to map the internal buffer ID (row number) to LOC_ID
loc_id_map <- data.frame(
  ID = 1:nrow(cam_1000m_buffer),  # This creates the internal IDs (1, 2, 3, ...)
  LOC_ID = cam_1000m_buffer$LOC_ID # And maps them to your LOC_ID attribute
)

# Join the LOC_ID to your 'vals' data.frame
lndc_vals<- merge(lndc_vals, loc_id_map, by = "ID")

# Group landcover classess into open, closed and unknown
lndc_vals_reclassified <- lndc_vals %>%
  mutate(
    landcover_class_grouped = case_when(
      `35K_20240101-20241231` %in% c(4, 5, 8, 11) ~ "open",
      `35K_20240101-20241231` %in% c(2, 3) ~ "closed",
      TRUE ~ "unknown" # Catches all other values
    )
  )

# Calculate proportions of different landcover classes within the buffer
lndc_prop <- lndc_vals_reclassified %>%
  group_by(LOC_ID, landcover_class_grouped) %>% # Group by each unique camera location and land cover class
  summarise(count = n(), .groups = "drop_last") %>% # Count the number of pixels for each combination
  group_by(LOC_ID) %>% # Now group only by camera location to calculate proportions within each buffer
  mutate(proportion = count / sum(count)) %>% # Calculate the proportion of each land cover class for that LOC_ID
  ungroup() %>%
  # *** NEW STEP: Select only the columns needed for pivoting to ensure 'count' is dropped ***
  dplyr::select(LOC_ID, landcover_class_grouped, proportion) %>%
  pivot_wider( # Reshape the data: make land cover classes into columns, fill with proportions
    names_from = landcover_class_grouped, # Values from this column become new column names
    values_from = proportion,           # Proportions fill the new columns
    values_fill = 0                     # Fill missing proportions with 0
  )

# Plot graph 
lndcpropplot <- ggplot(lndc_prop, aes(x = closed)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black") + # Adjust binwidth as needed
  labs(
    x = "Proportion of Closed Landcover",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Human population 

### Prepare raster - skip if reimporting merged raster in next step

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Worldpop/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("pop", i), rast(files[i]))
}

# Import country boundaries
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Mask by countries
pop_mask <- mask(pop1, study_countries)
pop_mask <- crop(pop_mask,study_countries)
plot(pop_mask)

# Re-projecting landcover rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
pop_aa <- project(pop_mask, africa_albers_crs, method = "near")

# Negiobourhood smoothing
pop_focal <- focal(pop_aa, w=101, fun=mean, na.rm=TRUE)
plot(pop_focal)

# Save reprojected merged raster
writeRaster(pop_aa, "popraster.tiff", overwrite=T)
```

### Calculate total within buffer

```{r}
# Read raster back in
pop_ras <- rast("C:/R Projects/Co-occurence/popraster.tiff") 

# Extract total values
popvals <- extract(pop_ras, cam_10km_buffer, "sum")

# Create a lookup table to map the internal buffer ID (row number) to LOC_ID
loc_id_map <- data.frame(
  ID = 1:nrow(cam_10km_buffer),  # This creates the internal IDs (1, 2, 3, ...)
  LOC_ID = cam_10km_buffer$LOC_ID # And maps them to your LOC_ID attribute
)

# Join the LOC_ID to your 'vals' data.frame
popvals<- merge(popvals, loc_id_map, by = "ID")

# Rename 2nd column
names(popvals)[2] <- "hum_pop"

# Log transform
popvals$log_hum_pop <- log1p(popvals$hum_pop)

# Plot graph
popplot <- ggplot(popvals, aes(x = log_hum_pop)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Log Transformed Total Population",
    y = "Count"
  ) +
  theme_minimal()
```

## Slope

### Prepare raster ~ Skip this step if reimporting just reimporting tiff file in next step

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Slope/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("slope", i), rast(files[i]))
}

# Re-projecting landcover rasters to AA ~ only including relevant TIFFs
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
slope1_aa <- project(slope1, africa_albers_crs, method = "near")
slope3_aa <- project(slope3, africa_albers_crs, method = "near")
slope4_aa <- project(slope4, africa_albers_crs, method = "near")
slope5_aa <- project(slope5, africa_albers_crs, method = "near")
slope7_aa <- project(slope7, africa_albers_crs, method = "near")
slope8_aa <- project(slope8, africa_albers_crs, method = "near")

# Merging landcover rasters
slope_aa <- merge(slope1_aa, slope3_aa, slope4_aa, slope5_aa, slope7_aa, slope8_aa)

# Check they match
plot(slope_aa)
plot(camlocs_vec, add=T)

writeRaster(slope_aa, "slope.tiff", overwrite=T)
```

### Caclulate average within buffer

```{r}
# Read raster back in
slope <- rast("C:/R Projects/Co-occurence/slope.tiff")

# Extract raster values inside each buffer polygon
slope_vals <- extract(slope, cam_1000m_buffer, fun = mean, na.rm = TRUE)

# Create a lookup table to map the internal buffer ID (row number) to LOC_ID
loc_id_map <- data.frame(
  ID = 1:nrow(cam_1000m_buffer),  # This creates the internal IDs (1, 2, 3, ...)
  LOC_ID = cam_1000m_buffer$LOC_ID # And maps them to your LOC_ID attribute
)

# Join the LOC_ID to your 'vals' data.frame
slope_vals<- merge(slope_vals, loc_id_map, by = "ID")

# Rename 2nd column
names(slope_vals)[2] <- "slope"

# Plot graph 
slopeplot <- ggplot(slope_vals, aes(x = slope)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") + 
  labs(
    x = "Average slope",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Roughness

### Prepare raster ~ Skip this step if reimporting just reimporting tiff file in next step

```{r}
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Roughness/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("dem", i), rast(files[i]))
}

# Re-projecting landcover rasters to AA 
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
dem1_aa <- project(dem1, africa_albers_crs, method = "near")
dem2_aa <- project(dem2, africa_albers_crs, method = "near")
dem3_aa <- project(dem3, africa_albers_crs, method = "near")

# Merging landcover rasters
dem_aa <- merge(dem1_aa, dem2_aa, dem3_aa)

# Check they match
plot(dem_aa)
plot(camlocs_vec, add=T)

writeRaster(dem_aa, "dem.tiff", overwrite=T)
```

### Calculate roughness within buffer

```{r}
# Read raster back in
elv <- rast("C:/R Projects/Co-occurence/dem.tiff")

# Calculate roughness within buffer
# crop & mask DEM to buffer extent
elv_crop <- mask(crop(elv, ext(cam_1000m_buffer)), cam_1000m_buffer)

# 3x3 focal window
w <- matrix(1, 3, 3)

# focal range (write to temp file to avoid memory issues)
range_r <- focal(
  elv_crop,
  w = w,
  fun = function(x, ...) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE) - min(x, na.rm = TRUE),
  na.policy = "omit",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

# extract mean roughness per buffer
res <- extract(range_r, cam_1000m_buffer, fun = mean, na.rm = TRUE)

# final output
rough_vals <- data.frame(
  LOC_ID = cam_1000m_buffer$LOC_ID,
  rough_range = res[[2]]
)

rough_vals

# Plot graph 
roughplot <- ggplot(rough_vals, aes(x = rough_range)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") + # Adjust binwidth as needed
  labs(
    x = "Average roughness",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Clay content 

### Prepare raster

```{r}
# read in raster
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/Soil/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("soil"), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting moisture rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
soil_aa <- project(soil, africa_albers_crs, method = "near")

# Check they match
plot(soil_aa)
plot(study_vect, add=T)

# Mask
soil_mask <- mask(soil_aa, study_vect)
soil_mask <- crop(soil_aa, study_vect)
soil_mask <- soil_mask[["mean_0_20"]]
plot(soil_mask)
plot(study_vect,add=T)

writeRaster(soil_mask, "soil_mask.tiff", overwrite=T)
```

### Calculate clay content wthin buffer

```{r}
# Read raster back in
soil <- rast("C:/R Projects/Co-occurence/soil_mask.tiff")

# Extract raster values inside each buffer polygon
soil_vals <- extract(soil, cam_1000m_buffer, fun = mean, na.rm = TRUE)

# Create a lookup table to map the internal buffer ID (row number) to LOC_ID
loc_id_map <- data.frame(
  ID = 1:nrow(cam_1000m_buffer),  # This creates the internal IDs (1, 2, 3, ...)
  LOC_ID = cam_1000m_buffer$LOC_ID # And maps them to your LOC_ID attribute
)

# Join the LOC_ID to your 'vals' data.frame
soil_vals<- merge(soil_vals, loc_id_map, by = "ID")

# Rename 2nd column
names(soil_vals)[2] <- "clay"

# Plot graph 
clayplot <- ggplot(soil_vals, aes(x = clay)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") + # Adjust binwidth as needed
  labs(
    x = "Average clay content",
    y = "Frequency"
  ) +
  theme_minimal()
```

# Add buffed covariates and write csv

```{r}
values <- merge(values, lndc_prop, by = "LOC_ID") 
values <- merge(values, popvals, by = "LOC_ID") 
values <- merge(values, slope_vals, by = "LOC_ID")
values <- merge(values, rough_vals, by = "LOC_ID")
values <- merge(values, soil_vals, by = "LOC_ID")

# update names for landcover
head(values)
names(values)[10] <- "closed_prop"
names(values)[11] <- "open_prop"
names(values)[12] <- "other_prop"
head(values)

spatial_covs <- as.data.frame(values)

spatial_covs <- spatial_covs[, !names(spatial_covs) %in% c("ID.x", "ID.y", "ID")]

write.csv(values, "spatial_covariates.csv", row.names=F)
```

## Plot covariate graphs

```{r}
# lndc
lndcplot <- ggplot(values, aes(x = land_cover)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Landcover",
    y = "Frequency"
  ) +
  theme_minimal()

# temp
tempplot <- ggplot(values, aes(x = temp)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Average temperature",
    y = "Frequency"
  ) +
  theme_minimal()

# temp seas
tempseasplot <- ggplot(values, aes(x = temp_seas)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Temperature seasonality",
    y = "Frequency"
  ) +
  theme_minimal()

# land temp
landtempplot <- ggplot(values, aes(x = land_temp)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Land surface temperature",
    y = "Frequency"
  ) +
  theme_minimal()

# precipitation
precipplot <- ggplot(values, aes(x = precip)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Average precipitation",
    y = "Frequency"
  ) +
  theme_minimal()

# precipitation seasonality
precipseasplot <- ggplot(values, aes(x = precip_seas)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Precipitation seasonality",
    y = "Frequency"
  ) +
  theme_minimal()

# distance to water
waterplot <- ggplot(values, aes(x = water_dist)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Distance to water",
    y = "Frequency"
  ) +
  theme_minimal()

# WDPA
wdpaplot <- ggplot(values, aes(x = WDPA)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(
    x = "Protected area classification",
    y = "Frequency"
  ) +
  theme_minimal()
```

# Run correleation test

```{r}
covs_num <- spatial_covs[, sapply(spatial_covs, is.numeric)]
cor_mat <- cor(covs_num, use = "complete.obs")

# coloured matrix plot
corr_plot <- corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.cex = 0.8,
  addCoef.col = "black",
  number.cex = 0.6
)
```



# Not required


## Soil moisture

### Prepare raster ~ skip if reimporting

```{r}
setwd("C:/Users/lady7240.BIOLOGY/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/Shapefiles/soil moisture/")

# List all TIFF files in the working directory
files <- list.files(pattern = "\\.tif$")

# Loop through each file and import
for (i in seq_along(files)) {
  assign(paste0("moist", i), rast(files[i]))
}

# Mask by country
# Download country outlines
countries <- ne_download(scale = "large", type = 'countries', returnclass="sf" )
study_countries <- countries[countries$NAME %in% c("Kenya", "Tanzania", "Zambia"),]

# Re-project countries into Africa Albers
africa_albers_crs <- st_crs("+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs")
study_countries_aea <- st_transform(study_countries, crs = africa_albers_crs)
study_vect <- vect(study_countries_aea)  # convert to terra format

# Re-projecting moisture rasters to AA
africa_albers_crs <- "+proj=aea +lat_1=-18 +lat_2=21 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
moist1_aa <- project(moist1, africa_albers_crs, method = "near")
moist2_aa <- project(moist2, africa_albers_crs, method = "near")
moist3_aa <- project(moist3, africa_albers_crs, method = "near")
moist4_aa <- project(moist4, africa_albers_crs, method = "near")
moist5_aa <- project(moist5, africa_albers_crs, method = "near")
moist6_aa <- project(moist6, africa_albers_crs, method = "near")

# Merging landcover rasters
moist_aa <- merge(moist1_aa, moist2_aa, moist3_aa, moist4_aa, moist5_aa, moist6_aa)

# Check they match
plot(moist_aa)
plot(study_vect, add=T)

# Mask
moist_mask <- mask(moist_aa, study_vect)
plot(moist_mask)
plot(study_vect,add=T)

writeRaster(moist_mask, "moist_mask.tiff", overwrite=T)
```

### Calculate average within radius

```{r}
# Read raster back in
moist_ras <- rast("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Analysis/Co-occurence/moist_mask.tiff") 

# Extract average values
moistvals <- extract(moist_ras, cam_1000m_buffer, "mean")

# Create a lookup table to map the internal buffer ID (row number) to LOC_ID
loc_id_map <- data.frame(
  ID = 1:nrow(cam_1000m_buffer),  # This creates the internal IDs (1, 2, 3, ...)
  LOC_ID = cam_1000m_buffer$LOC_ID # And maps them to your LOC_ID attribute
)

# Join the LOC_ID to your 'vals' data.frame
moistvals<- merge(moistvals, loc_id_map, by = "ID")

# Rename 2nd column
names(moistvals)[2] <- "soil_moist"

# Log transform
moistvals$log_soil_moist <- log1p(moistvals$soil_moist)

# Plot human population average
ggplot(moistvals, aes(x = log_soil_moist)) +
  geom_histogram(fill = "dodgerblue", color = "white") +
  labs(
    x = "Log Transformed Soil Moisture",
    y = "Count"
  ) +
  theme_minimal()
```
