---
title: "Co-occurence results plotting"
author: "Emily"
date: "2026-03-26"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load librarys

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(lubridate)
library(spOccupancy)
library(corrplot)
library(ggplot2)
library(viridis)
library(patchwork)
```

## Reload top model for further analysis

```{r}
top_model <- readRDS("top_model.rds")
occ_data <- readRDS("occ_data.rds")
```

# Posterior checks

```{r}
# Trace plots for convergence
plot(top_model, 'beta', density = FALSE)
summary(top_model)

# Perform a posterior predictive check to assess model fit.
#ppc.out <- ppcOcc(top_model, fit.stat = "freeman-tukey", group = 1)
#summary(ppc.out)

# The number of species in the data set
N <- nrow(top_model$y)
# Here I save the associated Bayesian p-value from each posterior predictive
# check for each species. You can save other information from the check, 
# but this is a small summary that won't take a lot of memory.
bayesian.p.vals.by.sp <- rep(NA, N)
mean.diff.sp <- rep(NA, N)
for (i in 1:N) {
  print(paste0("Currently on species ", i, " out of ", N))
  # Stuff to extract the detection random effects
  n.re.levels <- length(unique(top_model$X.p.re)) 
  re.indx <- ((i - 1) * n.re.levels + 1):(i * n.re.levels)
  # Generate the necessary object for ppcOcc for a single species at a time
  one.sp.out <- list(y = top_model$y[i, , ], # data set for the current species
		     z.samples = top_model$z.samples[, i, ], # latent occupancy values for the current species
		     n.post = top_model$n.post, # number of posterior samples for each chain
		     n.chains = top_model$n.chains, # number of chains
		     X.p = top_model$X.p, # the detection covariates
		     pRE = top_model$pRE, # Information on if there are any random effects
		     X.p.re = top_model$X.p.re, # Item for random effects
		     # Detection regression coefficient samples for current species only
		     alpha.samples = top_model$alpha.samples[, seq(i, ncol(top_model$alpha.samples), by = N)], 
         # Detection random effect variables
         alpha.star.samples = top_model$alpha.star.samples[, re.indx])
  # Set the class of one.sp.out, just to get get ppcOcc to work
  class(one.sp.out) <- 'PGOcc'
  # Calculate the posterior predictive check, see ?ppcOcc for different arguments
  # Note this step may take a few minutes for your data set.
  ppc.out <- ppcOcc(one.sp.out, fit.stat = 'freeman-tukey', group = 1)
  bayesian.p.vals.by.sp[i] <- mean(ppc.out$fit.y.rep > ppc.out$fit.y)
  mean.diff.sp[i] <- mean(ppc.out$fit.y.rep - ppc.out$fit.y)
  # Remove objects and release memory to help minimize RAM costs
  rm(ppc.out, one.sp.out)
  gc()
}
# Look at Bayesian p-values to assess model fit (want them to be close to 0.5)
bayesian.p.vals.by.sp
mean.diff.sp
```

# Community occurence plots

## Closed prop

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_closed <- top_model$beta.comm.samples[, "closed_prop"]

# Create closed prop values to predict from
closed_seq <- seq(
  min(occ_data$occ.covs$closed_prop, na.rm = TRUE),
  max(occ_data$occ.covs$closed_prop, na.rm = TRUE),
  length.out = 100)

n.samp <- length(beta0)
n.x <- length(closed_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)
for (i in 1:n.x) {
  eta <- beta0 + beta_closed * closed_seq[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_closed <- data.frame(
  closed_prop = closed_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_closed <- ggplot(df_closed, aes(x = closed_prop, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Proportion of closed habitat",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Terrain
```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_rough <- top_model$beta.comm.samples[, "scale(rough_range)"]

# Create gradient on ORIGINAL scale
rough_seq <- seq(
  min(occ_data$occ.covs$rough_range, na.rm = TRUE),
  max(occ_data$occ.covs$rough_range, na.rm = TRUE),
  length.out = 100
)

# Scale to match model
rough_scaled <- (rough_seq - mean(occ_data$occ.covs$rough_range, na.rm = TRUE)) /
                sd(occ_data$occ.covs$rough_range, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(rough_seq)

# Compute response
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_rough * rough_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_rough <- data.frame(
  rough = rough_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_rough <- ggplot(df_rough, aes(x = rough, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Terrain roughness",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Human pop

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_hum <- top_model$beta.comm.samples[, "scale(log_hum_pop)"]

# Create gradient on ORIGINAL scale
hum_seq <- seq(
  min(occ_data$occ.covs$log_hum_pop, na.rm = TRUE),
  max(occ_data$occ.covs$log_hum_pop, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
hum_scaled <- (hum_seq - mean(occ_data$occ.covs$log_hum_pop, na.rm = TRUE)) /
              sd(occ_data$occ.covs$log_hum_pop, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(hum_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_hum * hum_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_hum <- data.frame(
  log_hum_pop = hum_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_hum <- ggplot(df_hum, aes(x = log_hum_pop, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Log human population density",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## WDPA

```{r}
# Extract posterior samples
beta <- top_model$beta.comm.samples

# Get WDPA levels
wdpa_vals <- sort(unique(occ_data$occ.covs$WDPA))
wdpa_vals <- as.character(wdpa_vals)

# Identify WDPA coefficient columns
wdpa_cols <- grep("as.factor\\(WDPA\\)", colnames(beta), value = TRUE)

# Intercept
beta0 <- beta[, "(Intercept)"]

# Storage
n.samp <- nrow(beta)
n.wdpa <- length(wdpa_vals)

psi <- matrix(NA, nrow = n.wdpa, ncol = n.samp)

# Loop over WDPA levels
for (i in seq_along(wdpa_vals)) {
  level <- wdpa_vals[i]
  # baseline = intercept
  eta <- beta0
  # add effect if not reference level
  col_name <- paste0("as.factor(WDPA)", level)
  if (col_name %in% colnames(beta)) {
    eta <- eta + beta[, col_name]
  }
  psi[i, ] <- 1 / (1 + exp(-eta))
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_wdpa <- data.frame(
  WDPA = factor(wdpa_vals),
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_wdpa <- ggplot(df_wdpa, aes(x = WDPA, y = mean)) +
  geom_point(sze = 3, color = "#276221") +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.2, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Protected area category",
    y = "Mean community occupancy probability"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Rainfall

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_precip <- top_model$beta.comm.samples[, "scale(precip)"]

# Create gradient on ORIGINAL scale
precip_seq <- seq(
  min(occ_data$occ.covs$precip, na.rm = TRUE),
  max(occ_data$occ.covs$precip, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
precip_scaled <- (precip_seq - mean(occ_data$occ.covs$precip, na.rm = TRUE)) /
                 sd(occ_data$occ.covs$precip, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(precip_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_precip * precip_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_precip <- data.frame(
  precip = precip_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_precip <- ggplot(df_precip, aes(x = precip, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Average annual precipitation (mm)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Clay

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_clay <- top_model$beta.comm.samples[, "scale(clay)"]

# Create gradient on ORIGINAL scale
clay_seq <- seq(
  min(occ_data$occ.covs$clay, na.rm = TRUE),
  max(occ_data$occ.covs$clay, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
clay_scaled <- (clay_seq - mean(occ_data$occ.covs$clay, na.rm = TRUE)) /
               sd(occ_data$occ.covs$clay, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(clay_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_clay * clay_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_clay <- data.frame(
  clay = clay_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Plot
p_clay <- ggplot(df_clay, aes(x = clay, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Percentage clay content",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Water Distance

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_water <- top_model$beta.comm.samples[, "scale(water_dist)"]

# Create gradient on ORIGINAL scale
water_seq <- seq(
  min(occ_data$occ.covs$water_dist, na.rm = TRUE),
  max(occ_data$occ.covs$water_dist, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
water_scaled <- (water_seq - mean(occ_data$occ.covs$water_dist, na.rm = TRUE)) /
                sd(occ_data$occ.covs$water_dist, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(water_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_water * water_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_water <- data.frame(
  water_dist = water_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
) 

# Plot
p_water <- ggplot(df_water, aes(x = water_dist, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Distance to water (m)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Temperature

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_temp <- top_model$beta.comm.samples[, "scale(temp_seas)"]

# Create gradient on ORIGINAL scale
temp_seq <- seq(
  min(occ_data$occ.covs$temp_seas, na.rm = TRUE),
  max(occ_data$occ.covs$temp_seas, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
temp_scaled <- (temp_seq - mean(occ_data$occ.covs$temp_seas, na.rm = TRUE)) /
               sd(occ_data$occ.covs$temp_seas, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(temp_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_temp * temp_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_temp <- data.frame(
  temp_seas = temp_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Plot (x-axis on original scale)
p_temp <- ggplot(df_temp, aes(x = temp_seas, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Temperature seasonality",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Put plots all together

```{r}
# blank object for 9th grid
blank <- ggplot() +
  xlim(0, 1) +      
  ylim(0, 1) +     
  theme_classic() +
  theme(
    axis.text.y       = element_text(hjust = 1, face = "bold", colour = "white"),
    axis.text.x       = element_text(face = "bold", colour = "white"),
    axis.title.x      = element_text(face = "bold", colour = "white"),
    axis.title.y      = element_text(face = "bold", colour = "white"),
    axis.line         = element_line(colour = "white"),
    axis.ticks        = element_line(colour = "white"),
    panel.border      = element_blank(),
    panel.background  = element_rect(fill = "white", colour = NA),
    panel.grid        = element_blank(),
    plot.background   = element_rect(fill = "white", colour = NA)
  )

# plot together
all_plot <- (p_closed | p_rough | p_hum ) /
 (p_wdpa | p_temp | p_precip ) /
 (p_clay | p_water | blank)
all_plot
```

# Species association plots

## Species individual naive and predicted occupancy for comparison

```{r}
# --- Species occupancy table (model + naive) ---

z <- top_model$z.samples
y <- top_model$y

n.samp <- dim(z)[1]
n.sp   <- dim(z)[2]
n.site <- dim(z)[3]

sp.names <- dimnames(y)[[1]]

# --- Model-based occupancy ---
occ.sp <- matrix(NA, nrow = n.samp, ncol = n.sp)

for (s in 1:n.samp) {
  z_s <- z[s, , ]   # species × sites
  occ.sp[s, ] <- rowMeans(z_s)
}

occ.mean <- apply(occ.sp, 2, mean)
occ.low  <- apply(occ.sp, 2, quantile, 0.025)
occ.high <- apply(occ.sp, 2, quantile, 0.975)

# --- Naive occupancy ---
# detect if species observed at a site (any replicate = 1)
y_det <- apply(y, c(1, 2), max, na.rm = TRUE)  # species × site

naive_occ <- rowMeans(y_det)

# --- Combine into table ---
occ_table <- data.frame(
  species = sp.names,
  naive_occupancy = naive_occ,
  model_occupancy = occ.mean,
  low_95 = occ.low,
  high_95 = occ.high
)

# optional: sort
occ_table <- occ_table[order(-occ_table$model_occupancy), ]

occ_table
```


## Co-occurence plot

```{r}
# Extract dimensions
z <- top_model$z.samples
n.samp <- dim(z)[1]
n.sp   <- dim(z)[2]
n.site <- dim(z)[3]

# Create blank matrix
cooc.sum <- matrix(0, n.sp, n.sp)

# Calculate average co-occurence
for (s in 1:n.samp) {
  z_s <- matrix(z[s, , ], nrow = n.sp, ncol = n.site)
  # joint occurrence across sites
  cooc <- tcrossprod(z_s) / n.site
  cooc.sum <- cooc.sum + cooc
}

# Order species by size
size_order <- c(
  "Lion",
  "Leopard",
  "Spotted hyaena",
  "Striped hyaena",
  "African wild dog",
  "Cheetah",
  "Side-striped jackal",
  "Black-backed jackal",
  "Honey badger",
  "Caracal",
  "Serval",
  "Bat-eared fox",
  "Aardwolf",
  "African civet",
  "African wildcat"
)

# Mean co-occurence and attatch species names
cooc.mean <- cooc.sum / n.samp
rownames(cooc.mean) <- sp.names
colnames(cooc.mean) <- sp.names
cooc.mean <- cooc.mean[size_order, size_order]

# Plot
par(family = "sans", font = 2)  # bold labels

corrplot(
  cooc.mean,
  is.corr   = FALSE,          # not a correlation matrix (0–1 co‑occurrence)
  method    = "color",
  type      = "lower",
  diag      = FALSE,
  addCoef.col = "black",
  number.cex  = 0.6,
  tl.col      = "black",
  col.lim   = c(0, 1),        # because cooc.mean is 0–1
  col       = colorRampPalette(c("white", "#276221"))(200)
)
```

## Co-occurence accounting for low occupancy rates.

```{r}
# Create blank matrix
cooc.cond <- matrix(0, n.sp, n.sp)

# Calculate conditional co-occurrence P(B present | A present)
for (s in 1:n.samp) {
  z_s <- matrix(z[s, , ], nrow = n.sp, ncol = n.site)
  
  # For each species pair, count co-occurrences / occurrences of A
  for (i in 1:n.sp) {
    for (j in 1:n.sp) {
      if (i != j) {
        a_present <- sum(z_s[i, ])  # sites where A is present
        if (a_present > 0) {
          cooc.cond[i, j] <- cooc.cond[i, j] + sum(z_s[i, ] * z_s[j, ]) / a_present
        }
      }
    }
  }
}

# Mean conditional co-occurrence
cooc.cond.mean <- cooc.cond / n.samp
rownames(cooc.cond.mean) <- sp.names
colnames(cooc.cond.mean) <- sp.names
cooc.cond.mean <- cooc.cond.mean[size_order, size_order]

# Plot - full matrix now since it's asymmetric
par(family = "sans", font = 2)
corrplot(
  cooc.cond.mean,
  is.corr     = FALSE,
  method      = "color",
  type        = "full",
  diag        = FALSE,
  addCoef.col = "black",
  number.cex  = 0.6,
  tl.col      = "black",
  col.lim     = c(0, 1),
  col         = colorRampPalette(c("white", "#276221"))(200)
)
```


## Residual correlation matrix

```{r}
# Extract posterior latent‑rate samples (lambda.samples), then store the number of MCMC samples, the number of species, and the species names for downstream summaries or plots.
lam <- as.matrix(top_model$lambda.samples)
n.samp <- nrow(lam)
n.sp <- length(dimnames(top_model$y)[[1]])
sp.names <- dimnames(top_model$y)[[1]]

# infer number of factors
n.factors <- length(unique(as.integer(sub("^.*-", "", colnames(lam)))))

# create blank species by species matrix
corr.sum <- matrix(0, n.sp, n.sp)

# For each posterior MCMC sample, construct the factor‑loading matrix, compute the implied covariance matrix, convert it to a correlation matrix, and accumulate the correlation matrices across samples into a species‑by‑species correlation summary.
for (s in 1:n.samp) {
  L <- matrix(lam[s, ], nrow = n.sp, ncol = n.factors, byrow = TRUE)
  Sigma <- L %*% t(L)
  sd.vec <- sqrt(diag(Sigma))
  Corr <- Sigma / (sd.vec %o% sd.vec)
  corr.sum <- corr.sum + Corr
}

# Compute the mean species‑by‑species correlation matrix from the accumulated posterior correlation matrices, set the diagonal to 1 (perfect self‑correlation), and assign species names as both row and column labels for easier interpretation and plotting.
corr.mean <- corr.sum / n.samp
diag(corr.mean) <- 1
rownames(corr.mean) <- sp.names
colnames(corr.mean) <- sp.names
corr.mean <- corr.mean[size_order, size_order]

# Plot
par(family = "sans", font = 2)
corrplot(
  corr.mean,
  method = "color",
  type = "lower",
  diag = FALSE,
  addCoef.col = "black",
  number.cex = 0.6,
  tl.col = "black",
  col.lim = c(-1, 1),
  col = colorRampPalette(c("#40004B", "white", "#276221"))(200)
)
```

## Make network diagram

```{r}
library(ggplot2)
library(rphylopic)
library(scales)

# Extract all pairwise correlations (no threshold)
edge.mat <- which(upper.tri(corr.mean), arr.ind = TRUE)

edges <- data.frame(
  from = rownames(corr.mean)[edge.mat[,1]],
  to   = colnames(corr.mean)[edge.mat[,2]],
  corr = corr.mean[edge.mat]
)

# Create nodes and cluster ordering
nodes <- data.frame(name = rownames(corr.mean))
hc <- hclust(as.dist(1 - corr.mean))
nodes$name <- rownames(corr.mean)[hc$order]
n <- nrow(nodes)

# Circular layout
nodes$angle <- seq(0, 2*pi, length.out = n + 1)[-(n+1)]
nodes$x <- cos(nodes$angle)
nodes$y <- sin(nodes$angle)

# Label positions
nodes$x_lab <- cos(nodes$angle) * 1.05
nodes$y_lab <- sin(nodes$angle) * 1.05
nodes$x_lab[side_idx] <- cos(nodes$angle[side_idx]) * 1.15
nodes$y_lab[side_idx] <- sin(nodes$angle[side_idx]) * 1.15

# move up (increase y)
nodes$y_lab[nodes$name == "Cheetah"] <- nodes$y_lab[nodes$name == "Cheetah"] + 0.05
nodes$y_lab[nodes$name == "Serval"] <- nodes$y_lab[nodes$name == "Serval"] + 0.05
nodes$y_lab[nodes$name == "Lion"] <- nodes$y_lab[nodes$name == "Lion"] + 0.05
nodes$y_lab[nodes$name == "Aardwolf"] <- nodes$y_lab[nodes$name == "Aardwolf"] + 0.15

# move down (decrease y)
nodes$y_lab[nodes$name == "African wildcat"] <- nodes$y_lab[nodes$name == "African wildcat"] - 0.05

# move left
nodes$x_lab[nodes$name == "Spotted hyaena"] <- nodes$x_lab[nodes$name == "Spotted hyaena"] - 1.5

# Adjust label spacing for sides
top_bottom_idx <- abs(nodes$angle - pi/2) < pi/6 | abs(nodes$angle - 3*pi/2) < pi/6
side_idx <- !top_bottom_idx
side_offset <- 1.15
nodes$x_lab[side_idx] <- cos(nodes$angle[side_idx]) * side_offset
nodes$y_lab[side_idx] <- sin(nodes$angle[side_idx]) * side_offset

# UUID lookup (your correct mapping)
lookup <- data.frame(
  name = c(
    "Lion",
    "African wildcat",
    "African civet",
    "Leopard",
    "Side-striped jackal",
    "Striped hyaena",
    "Honey badger",
    "Black-backed jackal",
    "Aardwolf",
    "African wild dog",
    "Caracal",
    "Spotted hyaena",
    "Bat-eared fox",
    "Cheetah",
    "Serval"
  ),
  uuid = c(
    "de81a69c-3bdf-4486-b587-803e5b4acc64",
    "a3fb4098-a1e6-4f1c-9054-782ac8ab0eac",
    "cff1a00e-860b-4aaa-bdfa-eed4f917fb2b",
    "9f33e613-944c-483d-8275-cc59a1b470c2",
    "2bbc62f3-696f-4ac1-a235-6aeff17d2a88",
    "efd94e33-3966-4808-af47-3c58b333f9af",
    "8952c22d-1d0b-4658-b8ff-348df8aff18f",
    "94ab5d83-cdfe-4da8-b412-2528c843ffae",
    "e1072c6c-39fb-4375-aec5-dc0405ed1e3f",
    "832e5f72-4291-4916-bbbe-3a540b01fa1a",
    "0495f1b4-f9a0-4ba7-ae8f-bd89aa104c42",
    "f1b665ae-8fe9-42e4-b03a-4e9ae8213244",
    "39dc7ce6-3700-4d4a-9fdf-d84d8a0482e4",
    "0acb7026-d98d-4064-a009-0940580ea4b8",
    "01895510-7c27-489d-b441-6f72b1f562d2"
  )
)

# Match UUIDs without reordering
nodes$uuid <- lookup$uuid[match(nodes$name, lookup$name)]

# Adjust silhouette sizes
nodes$img_size <- 0.1
nodes$img_size[nodes$name == "African civet"] <- 0.06
nodes$img_size[nodes$name == "Honey badger"] <- 0.06

# Attach coordinates to edges
edges$x <- nodes$x[match(edges$from, nodes$name)]
edges$y <- nodes$y[match(edges$from, nodes$name)]
edges$xend <- nodes$x[match(edges$to, nodes$name)]
edges$yend <- nodes$y[match(edges$to, nodes$name)]

# Create curved edges
edge_curves <- do.call(rbind, lapply(1:nrow(edges), function(i) {
  x0 <- edges$x[i]
  y0 <- edges$y[i]
  x1 <- edges$xend[i]
  y1 <- edges$yend[i]
  cx <- (x0 + x1) / 2 * 0.6
  cy <- (y0 + y1) / 2 * 0.6
  t <- seq(0, 1, length.out = 50)
  data.frame(
    x = (1 - t)^2 * x0 + 2 * (1 - t) * t * cx + t^2 * x1,
    y = (1 - t)^2 * y0 + 2 * (1 - t) * t * cy + t^2 * y1,
    corr = as.numeric(edges$corr[i]),
    group = i
  )
}))

# Continuous linewidth scaling (fixes stepping issue)
edge_curves$width <- rescale(abs(edge_curves$corr), to = c(0.1, 2))

nodes[nodes$name == "Spotted hyaena", c("x_lab", "y_lab")]
nodes$x_lab[nodes$name == "Spotted hyaena"] <- -1.2

# Build plot
network <- ggplot() +
  geom_path(
    data = edge_curves,
    aes(x = x, y = y, group = group, colour = corr),
      linewidth = edge_curves$width,
      alpha = 0.9
  ) +
  geom_phylopic(
    data = nodes,
    aes(x = x, y = y, uuid = uuid, size = img_size),
    fill = "black"
  ) +
  scale_size_identity() +
  geom_text(
    data = nodes,
    aes(x = x_lab, y = y_lab, label = name),
    size = 3
  ) +
  scale_colour_gradient2(
    name = "Residual correlation",
    low = "#40004B",
    mid = "white",
    high = "#276221",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_equal(clip = "off") +
  guides(linewidth = "none") +
  theme_void()
```


## Alternative network

```{r}
library(igraph)
library(ggplot2)
library(scales)

# Build graph from your correlation matrix
corr.threshold <- 0.3

edge.mat <- which(corr.mean > corr.threshold & upper.tri(corr.mean), arr.ind = TRUE)

edges <- data.frame(
  from = rownames(corr.mean)[edge.mat[,1]],
  to   = colnames(corr.mean)[edge.mat[,2]],
  corr = corr.mean[edge.mat]
)

g <- graph_from_data_frame(
  edges,
  directed = FALSE,
  vertices = data.frame(name = rownames(corr.mean))
)

# Alignment
set.seed(123)
lay <- layout_with_fr(g)

deg <- degree(g)
main_idx <- which(deg > 0)
iso_idx  <- which(deg == 0)

# centre based on main cluster
center <- colMeans(lay[main_idx, ])

# radius of main cluster
main_radius <- max(sqrt(rowSums((lay[main_idx, ] - center)^2)))

# place isolates close to centre 
iso_radius <- main_radius * 0.6
angles <- seq(0, 2*pi, length.out = length(iso_idx) + 1)[-(length(iso_idx) + 1)]
lay[iso_idx, 1] <- center[1] + cos(angles) * iso_radius
lay[iso_idx, 2] <- center[2] + sin(angles) * iso_radius

# Extract node positions
nodes <- data.frame(
  name = V(g)$name,
  x = lay[,1],
  y = lay[,2]
)

# Add community detection (groups species with similar connections)
comm <- cluster_louvain(g)
nodes$community <- as.factor(comm$membership)

# community membership
membership <- comm$membership

# strength of pulling (0 = none, 1 = collapse)
pull_strength <- 0.3
for (k in unique(membership)) {
  idx <- which(membership == k)
  # centre of this community
  comm_center <- colMeans(lay[idx, , drop = FALSE])
  # move nodes toward their community centre
  lay[idx, ] <- lay[idx, ] * (1 - pull_strength) + 
                matrix(comm_center, nrow = length(idx), ncol = 2, byrow = TRUE) * pull_strength
}

# Attach coordinates to edges
edges$x <- nodes$x[match(edges$from, nodes$name)]
edges$y <- nodes$y[match(edges$from, nodes$name)]
edges$xend <- nodes$x[match(edges$to, nodes$name)]
edges$yend <- nodes$y[match(edges$to, nodes$name)]

network2 <- ggplot() +
  geom_segment(
    data = edges,
    aes(x = x, y = y, xend = xend, yend = yend, linewidth = corr),
    colour = "#276221",
    alpha = 0.6
  ) +
  geom_point(
    data = nodes,
    aes(x = x, y = y, fill = community),
    shape = 21,
    colour = "black",
    size = 4,
    stroke = 0.5
  ) +
  geom_text(
    data = nodes,
    aes(x = x, y = y, label = name),
    family = "Calibri",
    fontface = "bold",
    size = 3,
    vjust = -1
  ) +
  scale_linewidth(
    name = "Residual correlation (>0.5)",
    range = c(0.2, 6)
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_void(base_family = "Calibri")
```


# Make continuous plots

```{r}
p_temp_species   <- make_species_curve("temp_seas", "Temperature seasonality", scaled = TRUE)
p_precip_species <- make_species_curve("precip", "Average annual precipitation (mm)", scaled = TRUE)
p_clay_species   <- make_species_curve("clay", "Percentage clay content", scaled = TRUE)
p_water_species  <- make_species_curve("water_dist", "Distance to water (m)", scaled = TRUE)
p_closed_species <- make_species_curve("closed_prop", "Proportion of closed habitat", scaled = FALSE)
p_rough_species  <- make_species_curve("rough_range", "Terrain roughness", scaled = TRUE)
p_hum_species    <- make_species_curve("log_hum_pop", "Log human population density", scaled = TRUE)
```

## Plots protected area 

```{r}
beta <- as.matrix(top_model$beta.samples)
sp.names <- dimnames(top_model$y)[[1]]

# WDPA levels in the data
wdpa_vals <- sort(unique(occ_data$occ.covs$WDPA))

# Build dataframe of posterior mean occupancy by WDPA category for each species
curve_df <- do.call(rbind, lapply(sp.names, function(sp) {
  
  b0 <- beta[, paste0("(Intercept)-", sp)]
  
  wdpa_means <- sapply(wdpa_vals, function(w) {
    
    # baseline = intercept
    eta <- b0
    
    # add WDPA effect if not the reference category
    wdpa_term <- paste0("as.factor(WDPA)", w, "-", sp)
    
    if (wdpa_term %in% colnames(beta)) {
      eta <- eta + beta[, wdpa_term]
    }
    
    mean(plogis(eta))
  })
  
  data.frame(
    WDPA = factor(wdpa_vals, levels = wdpa_vals),
    mean = wdpa_means,
    species = sp
  )
}))

curve_df$species <- factor(curve_df$species, levels = sp.names)

# Plot
p_wdpa_species <- ggplot(curve_df, aes(x = WDPA, y = mean, color = species, linetype = species, group = species)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = grDevices::hcl.colors(length(sp.names), "Dark 3")) +
  scale_linetype_manual(values = rep(c("solid", "dashed", "dotted", "dotdash"),
                                     length.out = length(sp.names))) +
  labs(
    x = "Protected area category",
    y = "Occupancy probability",
    color = "Species",
    linetype = "Species"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank(),
    legend.position = "none"
  )

p_wdpa_species

# Remove legend completly
p_wdpa_species <- p_wdpa_species + guides(color = "none", linetype = "none")
```

## Combine

```{r}
blank <- plot_spacer()

allspecies_plot <-
  (p_closed_species + theme(axis.title.y = element_blank()) |
   p_rough_species  + theme(axis.title.y = element_blank()) |
   p_hum_species    + theme(axis.title.y = element_blank())) /
  (p_wdpa_species   + theme(axis.title.y = element_text()) |
   p_temp_species   + theme(axis.title.y = element_blank()) |
   p_precip_species + theme(axis.title.y = element_blank())) /
  (p_clay_species   + theme(axis.title.y = element_blank()) |
   p_water_species  + theme(axis.title.y = element_blank()) |
   blank) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

# Create table

```{r}
# Community + species coefficient table (one row per species)

fmt_ci <- function(est, low, high) {
  sprintf("%.2f [%.2f, %.2f]", est, low, high)
}

summarise_draws <- function(x) {
  x <- as.numeric(x)
  c(
    estimate = mean(x, na.rm = TRUE),
    lower_95 = as.numeric(quantile(x, 0.025, na.rm = TRUE)),
    upper_95 = as.numeric(quantile(x, 0.975, na.rm = TRUE))
  )
}

# Community-level samples
beta_comm <- as.matrix(top_model$beta.comm.samples)
comm_terms <- colnames(beta_comm)

comm_tab <- data.frame(
  row = "Community",
  stringsAsFactors = FALSE
)

for (term in comm_terms) {
  s <- summarise_draws(beta_comm[, term])
  comm_tab[[term]] <- fmt_ci(
    as.numeric(s["estimate"]),
    as.numeric(s["lower_95"]),
    as.numeric(s["upper_95"])
  )
}

# Species-level samples
beta_sp <- as.matrix(top_model$beta.samples)
sp.names <- dimnames(top_model$y)[[1]]

species_tabs <- lapply(sp.names, function(sp) {
  out <- data.frame(
    row = sp,
    stringsAsFactors = FALSE
  )
  
  for (term in comm_terms) {
    col_name <- paste0(term, "-", sp)
    if (col_name %in% colnames(beta_sp)) {
      s <- summarise_draws(beta_sp[, col_name])
      out[[term]] <- fmt_ci(
        as.numeric(s["estimate"]),
        as.numeric(s["lower_95"]),
        as.numeric(s["upper_95"])
      )
    } else {
      out[[term]] <- NA_character_
    }
  }
  
  out
})

# Combine community + species into one table
coef_table <- do.call(rbind, c(list(comm_tab), species_tabs))
rownames(coef_table) <- coef_table$row
coef_table$row <- NULL

coef_table
write.csv(coef_table, "coefficient_table.csv", row.names = TRUE)
```

## Calculate probability of directions

```{r}
posterior_direction <- function(x) {
  max(mean(x > 0, na.rm = TRUE), mean(x < 0, na.rm = TRUE))
}

make_pd_table <- function(mod, sp.names) {
  
  # community-level
  beta_comm <- as.matrix(mod$beta.comm.samples)
  comm_df <- tibble(
    taxon = "community",
    variable = colnames(beta_comm),
    pd = apply(beta_comm, 2, posterior_direction)
  )
  
  # species-level
  beta_sp <- as.matrix(mod$beta.samples)
  sp_df <- tibble(
    term = colnames(beta_sp),
    pd = apply(beta_sp, 2, posterior_direction),
    variable = sub("-.*$", "", term),
    taxon = sub("^[^-]+-", "", term)
  ) %>%
    select(taxon, variable, pd)
  
  # combine and reshape
  bind_rows(comm_df, sp_df) %>%
    mutate(
      taxon = factor(taxon, levels = c("community", sp.names))
    ) %>%
    arrange(taxon) %>%
    pivot_wider(names_from = variable, values_from = pd) %>%
    mutate(across(-taxon, ~ round(.x, 3)))
}

pd_table <- make_pd_table(top_model, sp.names)
pd_table
write.csv(pd_table, "pd_table.csv", row.names = TRUE)
```

## Full results table

```{r}
# One table with: beta, pd, and 95% CRI in each cell
# Format: beta, pd [lower, upper]

fmt_cell <- function(draws) {
  draws <- as.numeric(draws)
  
  est <- mean(draws, na.rm = TRUE)
  low <- as.numeric(quantile(draws, 0.025, na.rm = TRUE))
  high <- as.numeric(quantile(draws, 0.975, na.rm = TRUE))
  pd <- max(mean(draws > 0, na.rm = TRUE), mean(draws < 0, na.rm = TRUE))
  
  sprintf("%.2f, %.2f [%.2f, %.2f]", est, pd, low, high)
}

# Community + species coefficient table
make_full_table <- function(top_model, sp.names) {
  
  # Community-level samples
  beta_comm <- as.matrix(top_model$beta.comm.samples)
  comm_terms <- colnames(beta_comm)
  
  comm_tab <- data.frame(
    row = "Community",
    stringsAsFactors = FALSE
  )
  
  for (term in comm_terms) {
    comm_tab[[term]] <- fmt_cell(beta_comm[, term])
  }
  
  # Species-level samples
  beta_sp <- as.matrix(top_model$beta.samples)
  
  species_tabs <- lapply(sp.names, function(sp) {
    out <- data.frame(
      row = sp,
      stringsAsFactors = FALSE
    )
    
    for (term in comm_terms) {
      col_name <- paste0(term, "-", sp)
      
      if (col_name %in% colnames(beta_sp)) {
        out[[term]] <- fmt_cell(beta_sp[, col_name])
      } else {
        out[[term]] <- NA_character_
      }
    }
    
    out
  })
  
  # Combine community + species
  coef_table <- do.call(rbind, c(list(comm_tab), species_tabs))
  rownames(coef_table) <- coef_table$row
  coef_table$row <- NULL
  
  coef_table
}

# Build the table
sp.names <- dimnames(top_model$y)[[1]]
full_table <- make_full_table(top_model, sp.names)

full_table

# Save it
write.csv(full_table, "full_table.csv", row.names = TRUE)
```

