---
title: "Co-occurence main"
author: "Emily"
date: "2025-05-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(spOccupancy)
```

# Import raw station layout data files 

```{r}
#read data
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis")

# List all CSV files in the working directory
files <- list.files(pattern = "\\.csv$")

# Loop through each file and assign it based on its naming convention: -cap for capture data, -lay for station operation datess, -sp for station location
for (file in files) {
  # Extract base name without file extension and remove '_clean' part
  base <- tools::file_path_sans_ext(file)
  base <- sub("_clean", "", base)
  # Identify the type of file and assign it accordingly
  if (grepl("data", base)) {
    ds <- sub("data", "", base)
    assign(paste0(ds, "cap"), read.csv(file, header = TRUE))
  } else if (grepl("layout.+spatial", base)) {
    ds <- sub("layout", "", sub("_spatial", "", base))
    assign(paste0(ds, "sp"), read.csv(file, header = TRUE))
  } else if (grepl("layout", base)) {
    ds <- sub("layout", "", base)
    assign(paste0(ds, "lay"), read.csv(file, header = TRUE, check.names = FALSE))
  }
}
```

# Data preparation

## Formating DATE and LOC_ID columns

```{r}
# List of all objects
datasets <- ls()  

# Loop through each cap dataset to format date
for (ds in datasets) {
  if (grepl("cap$", ds)) {  # Select objects that name ends with "cap"
    df <- get(ds)
    df$DATE <- as.Date(df$DATE, format = "%Y-%m-%d")  # Convert DATE column to Date
    assign(ds, df, envir = .GlobalEnv)
  }
}

# Loop through each lay dataset to format date, remove blank rows/columns and add LOC-ID column name
for (ds in datasets) {
  if (grepl("lay$", ds)) { # Select objects that name ends with "lay"
    df <- get(ds)
    colnames(df) <- as.Date(colnames(df), format = "%d/%m/%Y") # Convert column names (dates)
    colnames(df)[1] <- "LOC_ID" # Name the first column LOC_ID
    keep_cols <- colSums(!(is.na(df) | df == "")) > 0  # Identify columns that are not empty
    df <- df[, keep_cols] # Drop columns that are completely empty
    keep_rows <- rowSums(!(is.na(df) | df == "")) > 0  # Identify rows that are not empty
    df <- df[keep_rows, ] # Drop rows that are completely empty
    assign(ds, df, envir = .GlobalEnv)
  }
}

# Loop through to read LOC_ID as character in layout files
for (ds in datasets) {
  if (grepl("lay$", ds)) {
  df <- get(ds)
  df$LOC_ID <- as.character(df$LOC_ID)
  assign(ds, df, envir = .GlobalEnv)
  }
}
```

## Shrink occasion length and calculate effort per week and create matrix for fieldseason variable

```{r}
for (ds in datasets) {
  if (grepl("cap$", ds)) { # Select cap files
    df_capture <- get(ds)
    df_capture <- df_capture %>%
    mutate(DATE = as.Date(DATE, format = "%Y-%m-%d"), WEEK = format(DATE, "%Y-%U")) # Use DATE to add WEEK column and values
    assign(ds, df_capture, envir = .GlobalEnv) # Overwrite the original
  }
  # Process layout datasets: create weekly status, effort, and fieldseason matrix
  if (grepl("lay$", ds)) {df_lay <- get(ds)
      lay_long <- df_lay %>%
      pivot_longer(-LOC_ID, names_to = "date", values_to = "status") %>%
      mutate(
        date = as.Date(date, format = "%Y-%m-%d"),
        WEEK = format(date, "%Y-%U")
      ) # Pivot to long
    
    # Weekly operational (1 if any day active)
    weekly <- lay_long %>%
      group_by(LOC_ID, WEEK) %>%
      summarise(lay_status = as.integer(any(status == 1, na.rm = TRUE)), .groups = "drop") %>%
      pivot_wider(names_from = WEEK, values_from = lay_status)
    
    # Weekly effort (count of active days)
    eff_weekly <- lay_long %>%
      group_by(LOC_ID, WEEK) %>%
      summarise(effort = sum(status == 1, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = WEEK, values_from = effort)
    
    # Fieldseason covariate matrix: replace 1s with file name (minus "lay")
    prefix <- sub("lay$", "", ds)
    lay_fs <- weekly
    lay_fs[,-1] <- lapply(lay_fs[,-1], function(col) ifelse(col == 1, prefix, NA))
    
    # Assign new objects back
    assign(paste0(prefix, "week"), weekly, envir = .GlobalEnv)
    assign(paste0(prefix, "eff"), eff_weekly, envir = .GlobalEnv)
    assign(paste0(prefix, "fs"), lay_fs, envir = .GlobalEnv)
  }
}
```

## Create detection histories

```{r}
#Prepare lists
week_names <- grep("week$", ls(), value = TRUE) #detection files
cap_names <- grep("cap$", ls(), value = TRUE) # capture files
site_list <- sort(unique(unlist(lapply(week_names, function(name) {
  get(name)$LOC_ID})))) #full list of sites
week_list <- sort(unique(unlist(lapply(week_names, function(name) {
  colnames(get(name))[-1]})))) # full list of weeks ([-1] excludes LOC_ID column)
species_list <- sort(unique(unlist(lapply(cap_names, function(name) {
  unique(get(name)$Species)})))) #full list of species

# Remove low detection species - edit as necessary
species_list <- setdiff(species_list, "Cheetah")

# Create empty array with lists
y_array <- array(NA, dim = c(length(species_list), length(site_list), length(week_list)),
                 dimnames = list(Species = species_list, Site = site_list, Week = week_list))

# Loop through each pair of cap and week datasets to fill y_array
for (i in seq_along(cap_names)) {
  cap_data <- get(cap_names[i])
  week_data <- get(week_names[i])
  for (w in seq_along(week_list)) {
    week_str <- week_list[w]
    # Subset capture data for current week
    cap_sub <- cap_data %>%
      filter(WEEK == week_str)
    # Subset layout data for current week
    if (!week_str %in% colnames(week_data)) next
    layout_sub <- week_data %>%
      select(LOC_ID, all_of(week_str)) %>%
      rename(op_status = 2)
    for (s in seq_along(site_list)) {
      site <- site_list[s]
      # Get operational status
      site_active <- layout_sub$op_status[layout_sub$LOC_ID == site]
      site_active <- ifelse(length(site_active) == 0, NA, site_active[1])
      for (sp in seq_along(species_list)) {
        species <- species_list[sp]
        if (!is.na(site_active) && site_active == 1) {
          detected <- any(cap_sub$LOC_ID == site & cap_sub$Species == species)
          y_array[sp, s, w] <- ifelse(detected, 1, 0)
        } else if (is.na(y_array[sp, s, w])) {
          y_array[sp, s, w] <- NA
        }
      }
    }
  }
}

# Calculate the number of total detections by week for each species
detection_totals <- apply(y_array, 1, function(x) sum(x, na.rm = TRUE))
detection_totals

# Reorder array
ordered_species <- names(sort(detection_totals, decreasing = TRUE))
species_index <- match(ordered_species, dimnames(y_array)[[1]])
y_array <- y_array[species_index,,]
#dimnames(y_array)[[3]] <- ordered_species
```

## Create spatial matrix (sites x coords)

```{r}
sp_list <- lapply(datasets[grepl("sp$", datasets)], get) # Select spatial dataframes
coords <- do.call(rbind, sp_list) # Merge 
coords <- coords[!duplicated(coords$LOC_ID), ] # Drop duplicate sites (From yearly repeat studies
coords <- coords[order(coords$LOC_ID), ] # Order alphabetically by LOC_ID
rownames(coords) <- coords$LOC_ID # Set LOC_ID as rownames
coords <- as.matrix(coords[, c("X_AfricaAlbers", "Y_AfricaAlbers")]) # Convert to a numeric matrix, using LOC_ID as rownames
coords <- coords[rownames(coords) %in% colnames(y_array), , drop = FALSE] # Remove rows which arent in y array
```

##  Create effort matrix (sites × weeks)

```{r}
eff_names <- grep("eff$", ls(), value = TRUE) # Identify all effort dataset names

# Pivot each to long form and stack them
eff_long <- lapply(eff_names, function(name) {
  df <- get(name)
  df %>%
    pivot_longer(-LOC_ID, names_to = "Week", values_to = "effort")
}) %>% bind_rows()

# Re-pivot to a single wide table, filling missing with NA
eff_merged <- eff_long %>%
  pivot_wider(
    names_from   = Week,
    values_from  = effort,
    values_fill  = NA
  )

# Sort rows (LOC_ID) alphabetically, and columns (Week) oldest→newest
eff_merged <- eff_merged %>%
  arrange(LOC_ID) %>%
  select(LOC_ID, sort(names(.)[-1]))  # keep LOC_ID first, then sorted weeks

# Convert to numeric matrix, dropping LOC_ID
effort_matrix <- as.matrix(eff_merged[ , -1, drop = FALSE])
rownames(effort_matrix) <- eff_merged$LOC_ID
effort_matrix[effort_matrix == 0] <- NA # Replace all 0s (no effort) with NA
storage.mode(effort_matrix) <- "double"
```

## Create field season matrix (sites × weeks)

```{r}
fs_names <- grep("fs$", ls(), value = TRUE) # Identify all fieldseason dataset names

# Pivot each to long form and stack them : LOC_ID | Week | fieldseason
fs_long <- lapply(fs_names, function(name) {
  df <- get(name)
  df %>%
    pivot_longer(-LOC_ID, names_to = "Week", values_to = "fieldseason")
}) %>% bind_rows()

# Re-pivot to a single wide table, filling missing with NA
fs_merged <- fs_long %>%
  pivot_wider(
    names_from   = Week,
    values_from  = fieldseason,
    values_fill  = NA
  )

# Sort rows (LOC_ID) alphabetically, and columns (Week) oldest→newest
fs_merged <- fs_merged %>%
  arrange(LOC_ID) %>%
  select(LOC_ID, sort(names(.)[-1]))  # keep LOC_ID first, then sorted weeks

# Convert to numeric matrix, dropping LOC_ID
fieldseason_matrix <- as.matrix(fs_merged[ , -1, drop = FALSE])
rownames(fieldseason_matrix) <- fs_merged$LOC_ID
storage.mode(fieldseason_matrix) <- "character"
fieldseason_matrix[fieldseason_matrix == "NULL"] <- NA
```
## Fieldseason as a numeric

```{r}
# Get unique non-NA values from the matrix
unique_vals <- na.omit(unique(as.vector(fieldseason_matrix)))

# Create a named vector mapping each unique string to a numeric code
val_map <- setNames(seq_along(unique_vals), unique_vals)

# Apply the mapping to the matrix
numeric_matrix <- matrix(
  data = val_map[as.character(fieldseason_matrix)],
  nrow = nrow(fieldseason_matrix),
  dimnames = dimnames(fieldseason_matrix)
)

# Convert to data.frame
fieldseason_numeric_matrix <- as.data.frame(numeric_matrix)
```

## Create landcover matrix for detection

```{r}
# Read data
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis")
spatial_covs <- read.csv("spatial_covariates.csv")

# Recode to bigger groups 
spatial_covs$land_cover_recoded <- ifelse(spatial_covs$land_cover %in% c(1,3,6,7,9,10), "unknown",
                                  ifelse(spatial_covs$land_cover == 2, "trees",
                                  ifelse(spatial_covs$land_cover %in% c(4,5,8,11), "rangeland",
                                         NA)))

# Create matrix
# Extract row names from effort_matrix
loc_ids <- rownames(effort_matrix)

# Lookup for recoded landcover per site
landcover_lookup <- setNames(spatial_covs$land_cover_recoded, spatial_covs$LOC_ID)

# Get landcover values per row
landcover_values <- landcover_lookup[loc_ids]

# Initialize an empty matrix with same dimensions and dimnames as effort_matrix
landcover_matrix <- matrix(NA_character_, 
                           nrow = nrow(effort_matrix), 
                           ncol = ncol(effort_matrix),
                           dimnames = dimnames(effort_matrix))

# Loop over rows or vectorize replacement for non-NA cells
for (i in seq_len(nrow(effort_matrix))) {
  # Get indices of cells that are not NA (or that have a value) in the current row
  valid_cols <- which(!is.na(effort_matrix[i, ]))  
  # Replace those cells with the landcover value for this site
  landcover_matrix[i, valid_cols] <- landcover_values[i]
}
```

# Covaraite correlation matrix

```{r}

```

# Create occupancy data list

```{r}
occ_data <- list(
  y = y_array,
  det.covs = list(
    effort = effort_matrix, 
    fieldseason = fieldseason_numeric_matrix,
    landcover = landcover_matrix
  ),
  coords = coords,
  occ.covs = spatial_covs
)
```

# Univariate selection

## Select detection probability covariates

## Select between protected area covariates

## Select between terrain covariates

## Select between temperature covariates

## Select between rainfall covariates

## Select between soil covariates

# Run multivariate models

```{r}
# Null model
null_model <- sfMsPGOcc(occ.formula = ~ 1,
                       det.formula = ~ 1,
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)
# Control model
det_model <- sfMsPGOcc(occ.formula = ~ scale(log_soil_moist),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 50,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)


# Combined model - dont forget to scale continuous variables
comb_model <- sfMsPGOcc(occ.formula = ~ closed_prop + scale(log_hum_pop) + scale(log_soil_moist),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)

# Habitat model
hab_model <- sfMsPGOcc(occ.formula = ~ closed_prop + scale(log_soil_moist),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)
# Human model
hum_model <- sfMsPGOcc(occ.formula = ~ scale(log_hum_pop) + scale(log_soil_moist),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)

```

## Posterior checks

```{r}
# Trace plots for convergence
plot(occ_model, 'beta', density = FALSE)

# Model validation --------------------------------------------------------
# Perform a posterior predictive check to assess model fit.
ppc.out <- ppcOcc(occ_model, fit.stat = 'freeman-tukey', group = 1)
# Calculate a Bayesian p-value as a simple measure of Goodness of Fit.
# Bayesian p-values between 0.1 and 0.9 indicate adequate model fit.
summary(ppc.out)

# Model comparison --------------------------------------------------------
# Compute Widely Applicable Information Criterion (WAIC)
# Lower values indicate better model fit.
waicOcc(hab_model, hum_model)

# Posterior summaries -----------------------------------------------------
# Concise summary of main parameter estimates
summary(occ_model, level = 'both')
```





# Trouble shooting

## Detection history issue - run per site to find where missing species data is

```{r}
# Prepare lists
week_names <- grep("SNNKLC2022week|SNNMSK2022week|SNNMTL2020week|SNSKPE2021week|SNSKPW2021week|SNSLWN2021week|SNSMGN2021week", ls(), value = TRUE) #detection files
cap_names <- grep("SNNKLC2022cap|SNNMSK2022cap|SNNMTL2020cap|SNSKPE2021cap|SNSKPW2021cap|SNSLWN2021cap|SNSMGN2021cap", ls(), value = TRUE) # capture files
site_list <- sort(unique(unlist(lapply(week_names, function(name) {
  get(name)$LOC_ID
})))) #full list of sites
week_list <- sort(unique(unlist(lapply(week_names, function(name) {
  colnames(get(name))[-1]  # Exclude LOC_ID column
})))) #full list of weeks
species_list <- sort(unique(unlist(lapply(cap_names, function(name) {
  unique(get(name)$Species)
})))) #full list of species


# Initialize empty array
y_array <- array(NA, dim = c(length(species_list), length(site_list), length(week_list)),
                 dimnames = list(Species = species_list, Site = site_list, Week = week_list))

# Fill the array from all datasets
# Loop through each pair of cap and week datasets
for (i in seq_along(cap_names)) {
  cap_data <- get(cap_names[i])
  week_data <- get(week_names[i])

  for (w in seq_along(week_list)) {
    week_str <- week_list[w]
    # Subset capture data for current week
    cap_sub <- cap_data %>%
      filter(WEEK == week_str)
    # Subset layout data for current week
    if (!week_str %in% colnames(week_data)) next
    layout_sub <- week_data %>%
      select(LOC_ID, all_of(week_str)) %>%
      rename(op_status = 2)
    for (s in seq_along(site_list)) {
      site <- site_list[s]
      # Get operational status
      site_active <- layout_sub$op_status[layout_sub$LOC_ID == site]
      site_active <- ifelse(length(site_active) == 0, NA, site_active[1])
      for (sp in seq_along(species_list)) {
        species <- species_list[sp]
        if (!is.na(site_active) && site_active == 1) {
          detected <- any(cap_sub$LOC_ID == site & cap_sub$Species == species)
          y_array[sp, s, w] <- ifelse(detected, 1, 0)
        } else if (is.na(y_array[sp, s, w])) {
          y_array[sp, s, w] <- NA
        }
      }
    }
  }
}

# Count detections
detection_totals <- apply(y_array, 1, function(x) sum(x, na.rm = TRUE))
detection_totals
```

## Error with some sites having empty detection histories

```{r}
d <- dim(y_array)
for (i in seq_along(d)) {
  cnt <- apply(y_array, i, function(slice) sum(!is.na(slice)))
  cat("\n--- Dimension", i, "(length =", d[i], ") ---\n")
  cat("slices with ANY non-NA:", sum(cnt > 0), "of", length(cnt), "\n")
  cat("first 20 counts:"); print(head(cnt, 20))
}


```

