---
title: "Co-occurence testing"
author: "Emily"
date: "2026-03-10"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(spOccupancy)
```

# Import raw station layout data files 

```{r}
#read data
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis")

# List all CSV files in the working directory
files <- list.files(pattern = "\\.csv$")

# Loop through each file and assign it based on its naming convention: -cap for capture data, -lay for station operation dates, -sp for station location
for (file in files) {
  # Extract base name without file extension and remove '_clean' part
  base <- tools::file_path_sans_ext(file)
  base <- sub("_clean", "", base)
  # Identify the type of file and assign it accordingly
  if (grepl("data", base)) {
    ds <- sub("data", "", base)
    assign(paste0(ds, "cap"), read.csv(file, header = TRUE))
  } else if (grepl("layout.+spatial", base)) {
    ds <- sub("layout", "", sub("_spatial", "", base))
    assign(paste0(ds, "sp"), read.csv(file, header = TRUE))
  } else if (grepl("layout", base)) {
    ds <- sub("layout", "", base)
    assign(paste0(ds, "lay"), read.csv(file, header = TRUE, check.names = FALSE))
  }
}

## TEMPORARY remove OJ files
rm(OJ2022cap, OJ2022sp)

```

# Data preparation

## Formating DATE and LOC_ID columns

```{r}
# List of all objects
datasets <- ls()  

# Loop through each cap dataset to format date
for (ds in datasets) {
  if (grepl("cap$", ds)) {  # Select objects that name ends with "cap"
    df <- get(ds)
    df$DATE <- as.Date(df$DATE, format = "%Y-%m-%d")  # Convert DATE column to Date
    assign(ds, df, envir = .GlobalEnv)
  }
}

# Loop through each lay dataset to format date, remove blank rows/columns and add LOC-ID column name
for (ds in datasets) {
  if (grepl("lay$", ds)) { # Select objects that name ends with "lay"
    df <- get(ds)
    colnames(df) <- as.Date(colnames(df), format = "%d/%m/%Y") # Convert column names (dates)
    colnames(df)[1] <- "LOC_ID" # Name the first column LOC_ID
    keep_cols <- colSums(!(is.na(df) | df == "")) > 0  # Identify columns that are not empty
    df <- df[, keep_cols] # Drop columns that are completely empty
    keep_rows <- rowSums(!(is.na(df) | df == "")) > 0  # Identify rows that are not empty
    df <- df[keep_rows, ] # Drop rows that are completely empty
    assign(ds, df, envir = .GlobalEnv)
  }
}

# Loop through to read LOC_ID as character in layout files
for (ds in datasets) {
  if (grepl("lay$", ds)) {
  df <- get(ds)
  df$LOC_ID <- as.character(df$LOC_ID)
  assign(ds, df, envir = .GlobalEnv)
  }
}
```

## Shrink occasion length and calculate effort per week and create matrix for fieldseason variable

```{r}
for (ds in datasets) {
  if (grepl("cap$", ds)) { # Select cap files
    df_capture <- get(ds)
    df_capture <- df_capture %>%
    mutate(DATE = as.Date(DATE, format = "%Y-%m-%d"), WEEK = format(DATE, "%Y-%W")) # Use DATE to add WEEK column and values
    assign(ds, df_capture, envir = .GlobalEnv) # Overwrite the original
  }
  # Process layout datasets: create weekly status, effort, and fieldseason matrix
  if (grepl("lay$", ds)) {df_lay <- get(ds)
      lay_long <- df_lay %>%
      pivot_longer(-LOC_ID, names_to = "date", values_to = "status") %>%
      mutate(
        date = as.Date(date, format = "%Y-%m-%d"),
        WEEK = format(date, "%Y-%W")
      ) # Pivot to long
    
    # Weekly operational (1 if any day active)
    weekly <- lay_long %>%
      group_by(LOC_ID, WEEK) %>%
      summarise(lay_status = as.integer(any(status == 1, na.rm = TRUE)), .groups = "drop") %>%
      pivot_wider(names_from = WEEK, values_from = lay_status)
    
    # Weekly effort (count of active days)
    eff_weekly <- lay_long %>%
      group_by(LOC_ID, WEEK) %>%
      summarise(effort = sum(status == 1, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = WEEK, values_from = effort)
    
    # Fieldseason covariate matrix: replace 1s with file name (minus "lay")
    prefix <- sub("lay$", "", ds) 
    lay_fs <- weekly 
    lay_fs[,-1] <- lapply(lay_fs[,-1], function(col) ifelse(col == 1, prefix, NA))
    
    # Assign new objects back
    assign(paste0(prefix, "week"), weekly, envir = .GlobalEnv)
    assign(paste0(prefix, "eff"), eff_weekly, envir = .GlobalEnv)
    assign(paste0(prefix, "fs"), lay_fs, envir = .GlobalEnv)
  }
}
```

## Calculate the number of detections per survey area per site

```{r}
# List all files ending in "cap"
files <- grep(pattern = "cap$", ls(), value = TRUE)

ZBRcap <- ZBRcap %>% mutate(across(c(LOC_ID), as.character))
ZBLLcap <- ZBLLcap %>% mutate(across(c(LOC_ID), as.character))

# Create an empty list to store results
results <- list()

# Loop through each data frame in the list
for (file_name in files) {
  df <- get(file_name)
  summary <- df |>
    dplyr::distinct(Species, WEEK, LOC_ID) |>
    dplyr::count(Species, name = "n_unique_week_loc")
  summary$file <- file_name
  results[[file_name]] <- summary
}


file_summary <- dplyr::bind_rows(results)

# Convert to wide table
file_summary_wide <- file_summary %>%
  pivot_wider(
    names_from = file,
    values_from = n_unique_week_loc,
    values_fill = 0  # fill missing species/file combos with 0
  )

file_summary_wide <- file_summary_wide %>%
  dplyr::mutate(TOTAL = rowSums(dplyr::across(where(is.numeric))))
```

## Create detection histories

```{r}
# Prepare lists
week_names <- grep("week$", ls(), value = TRUE) # detection files
cap_names <- grep("cap$", ls(), value = TRUE) # capture files
site_list <- sort(unique(unlist(lapply(week_names, function(name) {
  get(name)$LOC_ID})))) #full list of sites
week_list <- sort(unique(unlist(lapply(week_names, function(name) {
  colnames(get(name))[-1]})))) # full list of weeks ([-1] excludes LOC_ID column)
species_list <- sort(unique(unlist(lapply(cap_names, function(name) {
  unique(get(name)$Species)})))) #full list of species

# Remove low detection species - edit as necessary
# species_list <- setdiff(species_list, "Cheetah")

# Loop through cap files and merge
for (name in cap_names) {
  cap_data <- get(name)
  # Subset only the desired columns
  cap_data <- cap_data[, c("LOC_ID", "DATE", "Species", "WEEK")]
  # Overwrite the original data frame in the global environment
  assign(name, cap_data)
}
merged_caps <- dplyr::bind_rows(mget(cap_names))

## Make one big operational matrices
long_weeks <- lapply(week_names, function(name) {
  df <- get(name)
  df$LOC_ID <- as.character(df$LOC_ID)
  df_long <- pivot_longer(df, cols = -LOC_ID, names_to = "week", values_to = "op_status")
  return(df_long)
}) %>%
  bind_rows()

wide_weeks <- tidyr::pivot_wider(
  long_weeks,
  names_from = week,
  values_from = op_status
)

wide_weeks <- arrange(wide_weeks, LOC_ID)
week_cols <- setdiff(names(wide_weeks), "LOC_ID")
wide_weeks <- wide_weeks[, c("LOC_ID", sort(week_cols))]

wide_weeks[ , -1] <- lapply(wide_weeks[ , -1], function(col) {
  col <- ifelse(col == 0, NA, col)  # 0 → NA
  col <- ifelse(col == 1, 0, col)   # 1 → 0
  return(col)
})

# create y-array
site_list <- wide_weeks$LOC_ID
week_list <- setdiff(colnames(wide_weeks), "LOC_ID")

n_species <- length(species_list)
n_sites <- length(site_list)
n_weeks <- length(week_list)

y_array <- array(NA, dim = c(n_species, n_sites, n_weeks),
                 dimnames = list(Species = species_list, Site = site_list, Week = week_list))

week_matrix <- as.matrix(wide_weeks[ , week_list])

for (sp in seq_along(species_list)) {
  y_array[sp, , ] <- week_matrix
}

# populate y-array with species detections
na_attempts <- data.frame(Species = character(),
                          LOC_ID = character(),
                          WEEK = character(),
                          stringsAsFactors = FALSE)

for (i in seq_len(nrow(merged_caps))) {
  row <- merged_caps[i, ]
  Species <- as.character(row$Species)
  Site <- as.character(row$LOC_ID)
  Week <- as.character(row$WEEK)

  # Check if dimensions exist in y_array
  valid_species <- Species %in% dimnames(y_array)[[1]]
  valid_site    <- Site %in% dimnames(y_array)[[2]]
  valid_week    <- Week %in% dimnames(y_array)[[3]]
  if (valid_species && valid_site && valid_week) {
    # Safe to access the array
    current_value <- y_array[Species, Site, Week]
    if (is.na(current_value)) {
      # Exists but is NA — log it
      na_attempts <- rbind(na_attempts, data.frame(
        Species = Species,
        LOC_ID = Site,
        WEEK = Week,
        stringsAsFactors = FALSE
      ))
    } else {
      # Exists and is not NA — mark detection
      y_array[Species, Site, Week] <- 1
    }
  } else {
    # Combination doesn't exist in y_array — log it
    na_attempts <- rbind(na_attempts, data.frame(
      Species = Species,
      LOC_ID = Site,
      WEEK = Week,
      stringsAsFactors = FALSE
    ))
  }
}

# unique combinations in na_attempts
na_attempts <- na_attempts %>%
  distinct(Species, LOC_ID, WEEK, .keep_all = TRUE)




# Calculate the number of total detections by week for each species
detection_totals <- apply(y_array, 1, function(x) sum(x, na.rm = TRUE))
detection_totals

# Reorder array so that common species are first (better for multi species occupancy modelling)
#ordered_species <- names(sort(detection_totals, decreasing = TRUE))
#species_index <- match(ordered_species, dimnames(y_array)[[1]])
#y_array <- y_array[species_index,,]

# Temporarily remove Spotted hyaena from y_array
#y_array <- y_array[dimnames(y_array)[[1]] != "Spotted hyaena", , , drop = FALSE]
```

## Create spatial matrix (sites x coords)

```{r}
sp_list <- lapply(datasets[grepl("sp$", datasets)], get) # Select spatial dataframes
coords <- do.call(rbind, sp_list) # Merge 
coords <- coords[!duplicated(coords$LOC_ID), ] # Drop duplicate sites (From yearly repeat studies
coords <- coords[order(coords$LOC_ID), ] # Order alphabetically by LOC_ID
rownames(coords) <- coords$LOC_ID # Set LOC_ID as rownames
coords <- as.matrix(coords[, c("X_AfricaAlbers", "Y_AfricaAlbers")]) # Convert to a numeric matrix, using LOC_ID as rownames
coords <- coords[rownames(coords) %in% colnames(y_array), , drop = FALSE] # Remove rows which arent in y array
```

##  Create effort matrix (sites × weeks)

```{r}
eff_names <- grep("eff$", ls(), value = TRUE) # Identify all effort dataset names

# Pivot each to long form and stack them
eff_long <- lapply(eff_names, function(name) {
  df <- get(name)
  df %>%
    pivot_longer(-LOC_ID, names_to = "Week", values_to = "effort")
}) %>% bind_rows()

# Re-pivot to a single wide table, filling missing with NA
eff_merged <- eff_long %>%
  pivot_wider(
    names_from   = Week,
    values_from  = effort,
    values_fill  = NA
  )

# Sort rows (LOC_ID) alphabetically, and columns (Week) oldest→newest
eff_merged <- eff_merged %>%
  arrange(LOC_ID) %>%
  select(LOC_ID, sort(names(.)[-1]))  # keep LOC_ID first, then sorted weeks

# Convert to numeric matrix, dropping LOC_ID
effort_matrix <- as.matrix(eff_merged[ , -1, drop = FALSE])
rownames(effort_matrix) <- eff_merged$LOC_ID
effort_matrix[effort_matrix == 0] <- NA # Replace all 0s (no effort) with NA
storage.mode(effort_matrix) <- "double"
```

## Create field season matrix (sites × weeks) 

```{r}
fs_names <- setdiff(grep("fs$", ls(), value = TRUE), "lay_fs") # Identify all fieldseason dataset names

# Add column year to values
add_colname_to_cells_by_name <- function(nm) {
  if (!exists(nm, envir = .GlobalEnv)) return(invisible(NULL))
  mat <- get(nm, envir = .GlobalEnv)
  # require a matrix/data.frame with colnames and at least 2 columns
  if (is.null(dim(mat)) || is.null(colnames(mat)) || ncol(mat) < 2) {
    assign(nm, mat, envir = .GlobalEnv)
    return(invisible(NULL))
  }
  cols <- colnames(mat)[-1]   # ignore first column
  nr <- nrow(mat)
  for (col in cols) {
    for (r in seq_len(nr)) {
      base <- as.character(mat[r, col])        # single cell -> character
      mat[r, col] <- paste0(base, "_", col)    # assign single cell
    }
  }
  assign(nm, mat, envir = .GlobalEnv)
  invisible(NULL)
}

lapply(fs_names, add_colname_to_cells_by_name)

# reinstate NAs (may be a bit of a cicular way of doing this but i couldnt get the other ways to work)
set_NA_cells_by_name <- function(nm) {
  x <- get(nm, envir = .GlobalEnv)

  # operate element-by-element
  for (j in seq_len(ncol(x))) {
    for (i in seq_len(nrow(x))) {
      val <- x[i, j]
      if (!is.na(val) && grepl("NA", as.character(val))) {
        x[i, j] <- NA
      }
    }
  }

  assign(nm, x, envir = .GlobalEnv)
}

lapply(fs_names, set_NA_cells_by_name)

# remove week from names
remove_after_dash_by_name <- function(nm) {
  x <- get(nm, envir = .GlobalEnv)
  for (j in seq_len(ncol(x))) {
    for (i in seq_len(nrow(x))) {
      val <- x[i, j]
      if (!is.na(val)) {
        x[i, j] <- sub("-.*$", "", as.character(val))
      }
    }
  }
  assign(nm, x, envir = .GlobalEnv)
}

lapply(fs_names, remove_after_dash_by_name)

# Pivot each to long form and stack them : LOC_ID | Week | fieldseason
fs_long <- lapply(fs_names, function(name) {
  df <- get(name)
  df %>%
    pivot_longer(-LOC_ID, names_to = "Week", values_to = "fieldseason")
}) %>% bind_rows()

# Re-pivot to a single wide table, filling missing with NA
fs_merged <- fs_long %>%
  pivot_wider(
    names_from   = Week,
    values_from  = fieldseason,
    values_fill  = NA
  )

# Sort rows (LOC_ID) alphabetically, and columns (Week) oldest→newest
fs_merged <- fs_merged %>%
  arrange(LOC_ID) %>%
  select(LOC_ID, sort(names(.)[-1]))  # keep LOC_ID first, then sorted weeks

# Convert to numeric matrix, dropping LOC_ID
fieldseason_matrix <- as.matrix(fs_merged[ , -1, drop = FALSE])
rownames(fieldseason_matrix) <- fs_merged$LOC_ID
storage.mode(fieldseason_matrix) <- "character"
fieldseason_matrix[fieldseason_matrix == "NULL"] <- NA
```

## Fieldseason as a numeric

```{r}
# Get unique non-NA values from the matrix
unique_vals <- na.omit(unique(as.vector(fieldseason_matrix)))

# Create a named vector mapping each unique string to a numeric code
val_map <- setNames(seq_along(unique_vals), unique_vals)

# Apply the mapping to the matrix
numeric_matrix <- matrix(
  data = val_map[as.character(fieldseason_matrix)],
  nrow = nrow(fieldseason_matrix),
  dimnames = dimnames(fieldseason_matrix)
)

# Convert to data.frame
fieldseason_numeric_matrix <- as.data.frame(numeric_matrix)
```

## Create Targeted v's grid matrix

```{r}
# Copy the matrix structure
targetvgrid_matrix <- fieldseason_numeric_matrix

# Get row names
row_names <- rownames(fieldseason_numeric_matrix)

# Define prefix group that should get 2s
prefixes <- c("MN", "MT", "NB", "OMC")

# Loop over rows
for (i in seq_along(row_names)) {
  row_name <- row_names[i]
  # Check if row name starts with one of the target prefixes
  if (any(startsWith(row_name, prefixes))) {
    targetvgrid_matrix[i, !is.na(fieldseason_numeric_matrix[i, ])] <- 2
  } else {
    targetvgrid_matrix[i, !is.na(fieldseason_numeric_matrix[i, ])] <- 1
  }
}
```


## Create landcover matrix for detection

```{r}
# Read data
setwd("C:/Users/lady7240/OneDrive - Nexus365/Documents/3. Thesis/Chapter 3 - Co-occcurence/Data/For analysis")
spatial_covs <- read.csv("spatial_covariates - no OJ.csv", row.names=1)
spatial_covs <- spatial_covs[order(rownames(spatial_covs)), ]

# Recode to bigger groups 
spatial_covs$land_cover_recoded <- ifelse(spatial_covs$land_cover %in% c(1,3,6,7,9,10), "unknown",
                                  ifelse(spatial_covs$land_cover == 2, "trees",
                                  ifelse(spatial_covs$land_cover %in% c(4,5,8,11), "rangeland",
                                         NA)))

# Create matrix
# Extract row names from effort_matrix
loc_ids <- rownames(effort_matrix)

# Lookup for recoded landcover per site
landcover_lookup <- setNames(spatial_covs$land_cover_recoded, rownames(spatial_covs))

# Get landcover values per row
landcover_values <- landcover_lookup[loc_ids]

# Initialize an empty matrix with same dimensions and dimnames as effort_matrix
landcover_matrix <- matrix(NA_character_, 
                           nrow = nrow(effort_matrix), 
                           ncol = ncol(effort_matrix),
                           dimnames = dimnames(effort_matrix))

# Loop over rows or vectorize replacement for non-NA cells
for (i in seq_len(nrow(effort_matrix))) {
  # Get indices of cells that are not NA (or that have a value) in the current row
  valid_cols <- which(!is.na(effort_matrix[i, ]))  
  # Replace those cells with the landcover value for this site
  landcover_matrix[i, valid_cols] <- landcover_values[i]
}
```

## Fix missing WDPA values in spatial_covs for sites that are just outside shapefiles

```{r}
spatial_covs["KLKE_53", "WDPA"] <- 4
spatial_covs["MN09",    "WDPA"] <- 6
spatial_covs["NB06",    "WDPA"] <- 6
spatial_covs["NB49",    "WDPA"] <- 6
spatial_covs["WMA_01",  "WDPA"] <- 6
spatial_covs["WMA_02",  "WDPA"] <- 6
```

# Create occupancy data list

```{r}
occ_data <- list(
  y = y_array,
  det.covs = list(
    effort = effort_matrix, 
    fieldseason = fieldseason_numeric_matrix,
    landcover = landcover_matrix,
    targeted = targetvgrid_matrix
  ),
  coords = coords,
  occ.covs = spatial_covs
)

rm(list = setdiff(ls(envir = .GlobalEnv), "occ_data"), envir = .GlobalEnv)
```

# Univariate selection

## Run Null model

```{r}
# Null model
null_model <- msPGOcc(occ.formula = ~ 1,
                       det.formula = ~ 1,
                       data = occ_data,
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 5000,
                       n.samples = 10000,
                       n.burn = 3000,
                       n.thin = 2,
                       n.chains = 3)

plot(null_model, 'beta', density = FALSE)
summary(null_model)
```

## Select detection probability covariates

```{r}
# Full model
det_model1 <- msPGOcc(occ.formula = ~ 1,
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 5000,
                       n.samples = 45000,
                       n.burn = 3000,
                       n.thin = 4,
                       n.chains = 3)
summary(det_model1)

# Landcover
det_model2 <- msPGOcc(occ.formula = ~ 1,
                       det.formula = ~ effort  + as.factor(landcover) + (1|fieldseason), 
                       data = occ_data,
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 5000,
                       n.samples = 45000,
                       n.burn = 3000,
                       n.thin = 4,
                       n.chains = 3)
summary(det_model2)

# Targeted v's grid
det_model3 <- msPGOcc(occ.formula = ~ 1,
                       det.formula = ~ effort  + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 5000,
                       n.samples = 45000,
                       n.burn = 3000,
                       n.thin = 4,
                       n.chains = 3)
summary(det_model3)

# Core covariates
det_model4 <- msPGOcc(occ.formula = ~ 1,
                       det.formula = ~ effort + (1|fieldseason), 
                       data = occ_data,
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 5000,
                       n.samples = 45000,
                       n.burn = 3000,
                       n.thin = 4,
                       n.chains = 3)
summary(det_model4)

null_waic <- waicOcc(null_model)
det1_waic <- waicOcc(det_model1)
det2_waic <- waicOcc(det_model2)
det3_waic <- waicOcc(det_model3)
det4_waic <- waicOcc(det_model4)

null_waic
det1_waic
det2_waic
det3_waic
det4_waic
```

## Select between terrain covariates

```{r}
slope_model <- sfMsPGOcc(occ.formula = ~ scale(slope),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

rough_model <- sfMsPGOcc(occ.formula = ~ scale(rough_range),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

waicOcc(slope_model, rough_model)
```

## Select between temperature covariates

```{r}
temp_model <- sfMsPGOcc(occ.formula = ~ scale(temp),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

tempseas_model <- sfMsPGOcc(occ.formula = ~ scale(temp_seas),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

landtemp_model <- sfMsPGOcc(occ.formula = ~ scale(land_temp),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

waicOcc(temp_model, tempseas_model, landtemp_model)
```

## Select between rainfall covariates

```{r}
precip_model <- sfMsPGOcc(occ.formula = ~ scale(precip),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

precipseas_model <- sfMsPGOcc(occ.formula = ~ scale(precip_seas),
                       det.formula = ~ effort + as.factor(landcover) + as.factor(targeted) + (1|fieldseason), 
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 30,
                       batch.length = 100,
                       n.burn = 500,
                       n.thin = 5,
                       n.chains = 3)

waicOcc(precip_model, precipseas_model)
```


# Run multivariate models

```{r}
# Control model
cont_model <- sfMsPGOcc(occ.formula = ~ scale(temp) + scale(precip) + scale(clay) + scale(water_dist),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 50,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)

# Habitat model
hab_model <- sfMsPGOcc(occ.formula = ~ closed_prop, #scale(temp) + scale(precip) + scale(clay) + scale(water_dist) + scale (rough_range)
                       det.formula = ~ 1,# effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 200,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)


# Human model
hum_model <- sfMsPGOcc(occ.formula = ~ scale(temp) + scale(precip) + scale(clay) + scale(water_dist) + scale(log_hum_pop) + as.factor(WDPA),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)

# Combined model 
comb_model <- sfMsPGOcc(occ.formula = ~ scale(temp) + scale(precip) + scale(clay) + scale(water_dist) + closed_prop + scale (rough_range) + scale(log_hum_pop) + as.factor(WDPA),
                       det.formula = ~ effort + as.factor(landcover) + (1|fieldseason),
                       data = occ_data,
                       NNGP = TRUE,
                       n.neighbors = 10,
                       n.factors = 3,                      
                       verbose = TRUE,
                       n.omp.threads = 6,
                       n.report = 100,
                       n.batch = 400,
                       batch.length = 400,
                       n.burn = 5000,
                       n.thin = 5,
                       n.chains = 3)

waicOcc(null_model, det_model, cont_model, hab_model, hum_model, comb_model)
```

## Posterior checks

```{r}
# Trace plots for convergence
plot(det_model, 'beta', density = FALSE)

# Model validation --------------------------------------------------------
# Perform a posterior predictive check to assess model fit.
ppc.out <- ppcOcc(det_model, fit.stat = 'freeman-tukey', group = 1)

# Calculate a Bayesian p-value as a simple measure of Goodness of Fit.
# Bayesian p-values between 0.1 and 0.9 indicate adequate model fit.
summary(ppc.out)

# Model comparison --------------------------------------------------------
# Compute Widely Applicable Information Criterion (WAIC)
# Lower values indicate better model fit.
waicOcc(det_model, M1, m2)

# Posterior summaries -----------------------------------------------------
# Concise summary of main parameter estimates
summary(det_model, level = 'both')
```





# Trouble shooting

## Detection history issue - run per site to find where missing species data is

```{r}
# Prepare lists
week_names <- grep("SNNKLC2022week|SNNMSK2022week|SNNMTL2020week|SNSKPE2021week|SNSKPW2021week|SNSLWN2021week|SNSMGN2021week", ls(), value = TRUE) #detection files
cap_names <- grep("SNNKLC2022cap|SNNMSK2022cap|SNNMTL2020cap|SNSKPE2021cap|SNSKPW2021cap|SNSLWN2021cap|SNSMGN2021cap", ls(), value = TRUE) # capture files
site_list <- sort(unique(unlist(lapply(week_names, function(name) {
  get(name)$LOC_ID
})))) #full list of sites
week_list <- sort(unique(unlist(lapply(week_names, function(name) {
  colnames(get(name))[-1]  # Exclude LOC_ID column
})))) #full list of weeks
species_list <- sort(unique(unlist(lapply(cap_names, function(name) {
  unique(get(name)$Species)
})))) #full list of species


# Initialize empty array
y_array <- array(NA, dim = c(length(species_list), length(site_list), length(week_list)),
                 dimnames = list(Species = species_list, Site = site_list, Week = week_list))

# Fill the array from all datasets
# Loop through each pair of cap and week datasets
for (i in seq_along(cap_names)) {
  cap_data <- get(cap_names[i])
  week_data <- get(week_names[i])

  for (w in seq_along(week_list)) {
    week_str <- week_list[w]
    # Subset capture data for current week
    cap_sub <- cap_data %>%
      filter(WEEK == week_str)
    # Subset layout data for current week
    if (!week_str %in% colnames(week_data)) next
    layout_sub <- week_data %>%
      select(LOC_ID, all_of(week_str)) %>%
      rename(op_status = 2)
    for (s in seq_along(site_list)) {
      site <- site_list[s]
      # Get operational status
      site_active <- layout_sub$op_status[layout_sub$LOC_ID == site]
      site_active <- ifelse(length(site_active) == 0, NA, site_active[1])
      for (sp in seq_along(species_list)) {
        species <- species_list[sp]
        if (!is.na(site_active) && site_active == 1) {
          detected <- any(cap_sub$LOC_ID == site & cap_sub$Species == species)
          y_array[sp, s, w] <- ifelse(detected, 1, 0)
        } else if (is.na(y_array[sp, s, w])) {
          y_array[sp, s, w] <- NA
        }
      }
    }
  }
}

# Count detections
detection_totals <- apply(y_array, 1, function(x) sum(x, na.rm = TRUE))
detection_totals
```

## Error with some sites having empty detection histories

```{r}
d <- dim(y_array)
for (i in seq_along(d)) {
  cnt <- apply(y_array, i, function(slice) sum(!is.na(slice)))
  cat("\n--- Dimension", i, "(length =", d[i], ") ---\n")
  cat("slices with ANY non-NA:", sum(cnt > 0), "of", length(cnt), "\n")
  cat("first 20 counts:"); print(head(cnt, 20))
}


```

## Finding missing sites

```{r}
na_sites <- which(!complete.cases(occ_data$occ.covs))

site_na_report <- lapply(na_sites, function(i) {
  colnames(occ_data$occ.covs)[is.na(occ_data$occ.covs[i, ])]
})

names(site_na_report) <- rownames(occ_data$occ.covs)[na_sites]

site_na_report
```

## Find missing det.covs

```{r}
y <- occ_data$y          # species x sites x visits (array) OR sites x visits (matrix)
det.covs <- occ_data$det.covs
observed_y <- apply(!is.na(y), c(2,3), any)

cov_names <- names(det.covs)

conflict_list <- lapply(cov_names, function(cn) {
  cov_mat <- det.covs[[cn]]
  
  # Ensure shape matches y
  if (!all(dim(cov_mat) == dim(observed_y))) {
    stop(paste("Dimension mismatch for covariate:", cn))
  }
  
  # Conflict = cov is NA AND y has real observation
  conflicts <- is.na(cov_mat) & observed_y
  
  if (any(conflicts)) {
    which(conflicts, arr.ind = TRUE)
  } else {
    NULL
  }
})

names(conflict_list) <- cov_names
conflict_list

```

