---
title: "Co-occurence results plotting"
author: "Emily"
date: "2026-03-26"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(spOccupancy)
library(corrplot)
library(ggplot2)
library(viridis)
library(patchwork)
```

## Reload top model for further analysis

```{r}
top_model <- readRDS("top_model.rds")
occ_data <- readRDS("occ_data.rds")
```

## Posterior checks

```{r}
# Trace plots for convergence
plot(top_model, 'beta', density = FALSE)

# Model validation --------------------------------------------------------
# Perform a posterior predictive check to assess model fit.
ppc.out <- ppcOcc(top_model, fit.stat = "chi-squared", group = 1)
summary(ppc.out)

# Posterior summaries -----------------------------------------------------
# Concise summary of main parameter estimates
summary(top_model, level = 'both')
```

# Community occurence plots

## Closed prop

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_closed <- top_model$beta.comm.samples[, "closed_prop"]

# Create closed prop values to predict from
closed_seq <- seq(
  min(occ_data$occ.covs$closed_prop, na.rm = TRUE),
  max(occ_data$occ.covs$closed_prop, na.rm = TRUE),
  length.out = 100)

n.samp <- length(beta0)
n.x <- length(closed_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)
for (i in 1:n.x) {
  eta <- beta0 + beta_closed * closed_seq[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_closed <- data.frame(
  closed_prop = closed_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_closed <- ggplot(df_closed, aes(x = closed_prop, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Proportion of closed habitat",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Terrain
```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_rough <- top_model$beta.comm.samples[, "scale(rough_range)"]

# Create gradient on ORIGINAL scale
rough_seq <- seq(
  min(occ_data$occ.covs$rough_range, na.rm = TRUE),
  max(occ_data$occ.covs$rough_range, na.rm = TRUE),
  length.out = 100
)

# Scale to match model
rough_scaled <- (rough_seq - mean(occ_data$occ.covs$rough_range, na.rm = TRUE)) /
                sd(occ_data$occ.covs$rough_range, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(rough_seq)

# Compute response
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_rough * rough_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_rough <- data.frame(
  rough = rough_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_rough <- ggplot(df_rough, aes(x = rough, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Terrain roughness",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Human pop

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_hum <- top_model$beta.comm.samples[, "scale(log_hum_pop)"]

# Create gradient on ORIGINAL scale
hum_seq <- seq(
  min(occ_data$occ.covs$log_hum_pop, na.rm = TRUE),
  max(occ_data$occ.covs$log_hum_pop, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
hum_scaled <- (hum_seq - mean(occ_data$occ.covs$log_hum_pop, na.rm = TRUE)) /
              sd(occ_data$occ.covs$log_hum_pop, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(hum_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_hum * hum_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

# Combine into dataframe
df_hum <- data.frame(
  log_hum_pop = hum_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_hum <- ggplot(df_hum, aes(x = log_hum_pop, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Log human population density",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## WDPA

```{r}
# Extract posterior samples
beta <- top_model$beta.comm.samples

# Get WDPA levels
wdpa_vals <- sort(unique(occ_data$occ.covs$WDPA))
wdpa_vals <- as.character(wdpa_vals)

# Identify WDPA coefficient columns
wdpa_cols <- grep("as.factor\\(WDPA\\)", colnames(beta), value = TRUE)

# Intercept
beta0 <- beta[, "(Intercept)"]

# Storage
n.samp <- nrow(beta)
n.wdpa <- length(wdpa_vals)

psi <- matrix(NA, nrow = n.wdpa, ncol = n.samp)

# Loop over WDPA levels
for (i in seq_along(wdpa_vals)) {
  level <- wdpa_vals[i]
  # baseline = intercept
  eta <- beta0
  # add effect if not reference level
  col_name <- paste0("as.factor(WDPA)", level)
  if (col_name %in% colnames(beta)) {
    eta <- eta + beta[, col_name]
  }
  psi[i, ] <- 1 / (1 + exp(-eta))
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_wdpa <- data.frame(
  WDPA = factor(wdpa_vals),
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_wdpa <- ggplot(df_wdpa, aes(x = WDPA, y = mean)) +
  geom_point(sze = 3, color = "#276221") +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.2, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Protected area category",
    y = "Mean community occupancy probability"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Rainfall

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_precip <- top_model$beta.comm.samples[, "scale(precip)"]

# Create gradient on ORIGINAL scale
precip_seq <- seq(
  min(occ_data$occ.covs$precip, na.rm = TRUE),
  max(occ_data$occ.covs$precip, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
precip_scaled <- (precip_seq - mean(occ_data$occ.covs$precip, na.rm = TRUE)) /
                 sd(occ_data$occ.covs$precip, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(precip_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_precip * precip_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_precip <- data.frame(
  precip = precip_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Create plot
p_precip <- ggplot(df_precip, aes(x = precip, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Average annual precipitation (mm)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Clay

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_clay <- top_model$beta.comm.samples[, "scale(clay)"]

# Create gradient on ORIGINAL scale
clay_seq <- seq(
  min(occ_data$occ.covs$clay, na.rm = TRUE),
  max(occ_data$occ.covs$clay, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
clay_scaled <- (clay_seq - mean(occ_data$occ.covs$clay, na.rm = TRUE)) /
               sd(occ_data$occ.covs$clay, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(clay_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_clay * clay_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_clay <- data.frame(
  clay = clay_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Plot
p_clay <- ggplot(df_clay, aes(x = clay, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Percentage clay content",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Water Distance

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_water <- top_model$beta.comm.samples[, "scale(water_dist)"]

# Create gradient on ORIGINAL scale
water_seq <- seq(
  min(occ_data$occ.covs$water_dist, na.rm = TRUE),
  max(occ_data$occ.covs$water_dist, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
water_scaled <- (water_seq - mean(occ_data$occ.covs$water_dist, na.rm = TRUE)) /
                sd(occ_data$occ.covs$water_dist, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(water_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_water * water_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_water <- data.frame(
  water_dist = water_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
) 

# Plot
p_water <- ggplot(df_water, aes(x = water_dist, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Distance to water (m)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Temperature

```{r}
# Extract posterior samples
beta0 <- top_model$beta.comm.samples[, "(Intercept)"]
beta_temp <- top_model$beta.comm.samples[, "scale(temp_seas)"]

# Create gradient on ORIGINAL scale
temp_seq <- seq(
  min(occ_data$occ.covs$temp_seas, na.rm = TRUE),
  max(occ_data$occ.covs$temp_seas, na.rm = TRUE),
  length.out = 100
)

# Scale using original data (to match model)
temp_scaled <- (temp_seq - mean(occ_data$occ.covs$temp_seas, na.rm = TRUE)) /
               sd(occ_data$occ.covs$temp_seas, na.rm = TRUE)

n.samp <- length(beta0)
n.x <- length(temp_seq)

# Compute response curve
psi <- matrix(NA, nrow = n.x, ncol = n.samp)

for (i in 1:n.x) {
  eta <- beta0 + beta_temp * temp_scaled[i]
  psi[i, ] <- 1 / (1 + exp(-eta))   # inverse logit
}

# Summarise posterior
psi.mean <- apply(psi, 1, mean)
psi.low  <- apply(psi, 1, quantile, 0.025)
psi.high <- apply(psi, 1, quantile, 0.975)

df_temp <- data.frame(
  temp_seas = temp_seq,
  mean = psi.mean,
  low = psi.low,
  high = psi.high
)

# Plot (x-axis on original scale)
p_temp <- ggplot(df_temp, aes(x = temp_seas, y = mean)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = "#276221", alpha = 0.2) +
  geom_line(linewidth = 1, color = "#276221") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Temperature seasonality",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank()
  )
```

## Put plots all together

```{r}
# blank object for 9th grid
blank <- ggplot() +
  xlim(0, 1) +      
  ylim(0, 1) +     
  theme_classic() +
  theme(
    axis.text.y       = element_text(hjust = 1, face = "bold", colour = "white"),
    axis.text.x       = element_text(face = "bold", colour = "white"),
    axis.title.x      = element_text(face = "bold", colour = "white"),
    axis.title.y      = element_text(face = "bold", colour = "white"),
    axis.line         = element_line(colour = "white"),
    axis.ticks        = element_line(colour = "white"),
    panel.border      = element_blank(),
    panel.background  = element_rect(fill = "white", colour = NA),
    panel.grid        = element_blank(),
    plot.background   = element_rect(fill = "white", colour = NA)
  )

# plot together
all_plot <- (p_closed | p_rough | p_hum ) /
 (p_wdpa | p_temp | p_precip ) /
 (p_clay | p_water | blank)
all_plot
```

# Species association plots

## Species individual naive and predicted occupancy for comparison

```{r}
# --- Species occupancy table (model + naive) ---

z <- top_model$z.samples
y <- top_model$y

n.samp <- dim(z)[1]
n.sp   <- dim(z)[2]
n.site <- dim(z)[3]

sp.names <- dimnames(y)[[1]]

# --- Model-based occupancy ---
occ.sp <- matrix(NA, nrow = n.samp, ncol = n.sp)

for (s in 1:n.samp) {
  z_s <- z[s, , ]   # species × sites
  occ.sp[s, ] <- rowMeans(z_s)
}

occ.mean <- apply(occ.sp, 2, mean)
occ.low  <- apply(occ.sp, 2, quantile, 0.025)
occ.high <- apply(occ.sp, 2, quantile, 0.975)

# --- Naive occupancy ---
# detect if species observed at a site (any replicate = 1)
y_det <- apply(y, c(1, 2), max, na.rm = TRUE)  # species × site

naive_occ <- rowMeans(y_det)

# --- Combine into table ---
occ_table <- data.frame(
  species = sp.names,
  naive_occupancy = naive_occ,
  model_occupancy = occ.mean,
  low_95 = occ.low,
  high_95 = occ.high
)

# optional: sort
occ_table <- occ_table[order(-occ_table$model_occupancy), ]

occ_table
```


## Co-occurence plot

```{r}
# Extract dimensions
z <- top_model$z.samples
n.samp <- dim(z)[1]
n.sp   <- dim(z)[2]
n.site <- dim(z)[3]

# Create blank matrix
cooc.sum <- matrix(0, n.sp, n.sp)

# Calculate average co-occurence
for (s in 1:n.samp) {
  z_s <- matrix(z[s, , ], nrow = n.sp, ncol = n.site)
  # joint occurrence across sites
  cooc <- tcrossprod(z_s) / n.site
  cooc.sum <- cooc.sum + cooc
}

# Order species by size
size_order <- c(
  "Lion",
  "Leopard",
  "Spotted hyaena",
  "Striped hyaena",
  "African wild dog",
  "Cheetah",
  "Side-striped jackal",
  "Black-backed jackal",
  "Honey badger",
  "Caracal",
  "Serval",
  "Bat-eared fox",
  "Aardwolf",
  "African civet",
  "African wildcat"
)

# Mean co-occurence and attatch species names
cooc.mean <- cooc.sum / n.samp
rownames(cooc.mean) <- sp.names
colnames(cooc.mean) <- sp.names
cooc.mean <- cooc.mean[size_order, size_order]

# Plot
par(family = "sans", font = 2)  # bold labels

corrplot(
  cooc.mean,
  is.corr   = FALSE,          # not a correlation matrix (0–1 co‑occurrence)
  method    = "color",
  type      = "lower",
  diag      = FALSE,
  addCoef.col = "black",
  number.cex  = 0.6,
  tl.col      = "black",
  col.lim   = c(0, 1),        # because cooc.mean is 0–1
  col       = colorRampPalette(c("white", "#276221"))(200)
)
```

## Residual correlation matrix

```{r}
# Extract posterior latent‑rate samples (lambda.samples), then store the number of MCMC samples, the number of species, and the species names for downstream summaries or plots.
lam <- as.matrix(top_model$lambda.samples)
n.samp <- nrow(lam)
n.sp <- length(dimnames(top_model$y)[[1]])
sp.names <- dimnames(top_model$y)[[1]]

# infer number of factors
n.factors <- length(unique(as.integer(sub("^.*-", "", colnames(lam)))))

# create blank species by species matrix
corr.sum <- matrix(0, n.sp, n.sp)

# For each posterior MCMC sample, construct the factor‑loading matrix, compute the implied covariance matrix, convert it to a correlation matrix, and accumulate the correlation matrices across samples into a species‑by‑species correlation summary.
for (s in 1:n.samp) {
  L <- matrix(lam[s, ], nrow = n.sp, ncol = n.factors, byrow = TRUE)
  Sigma <- L %*% t(L)
  sd.vec <- sqrt(diag(Sigma))
  Corr <- Sigma / (sd.vec %o% sd.vec)
  corr.sum <- corr.sum + Corr
}

# Compute the mean species‑by‑species correlation matrix from the accumulated posterior correlation matrices, set the diagonal to 1 (perfect self‑correlation), and assign species names as both row and column labels for easier interpretation and plotting.
corr.mean <- corr.sum / n.samp
diag(corr.mean) <- 1
rownames(corr.mean) <- sp.names
colnames(corr.mean) <- sp.names
corr.mean <- corr.mean[size_order, size_order]

# Plot
par(family = "sans", font = 2)
corrplot(
  corr.mean,
  method = "color",
  type = "lower",
  diag = FALSE,
  addCoef.col = "black",
  number.cex = 0.6,
  tl.col = "black",
  col.lim = c(-1, 1),
  col = colorRampPalette(c("#c30000", "white", "#276221"))(200)
)
```

# Individual species response curves

## Create function to make species plots

```{r}
make_species_curve <- function(focal_var, xlab_text, scaled = FALSE) {
  
  beta <- as.matrix(top_model$beta.samples)
  sp.names <- dimnames(top_model$y)[[1]]
  
  # original-scale gradient for plotting
  x_seq <- seq(
    min(occ_data$occ.covs[[focal_var]], na.rm = TRUE),
    max(occ_data$occ.covs[[focal_var]], na.rm = TRUE),
    length.out = 100
  )
  
  # scale if the variable was scaled in the model
  if (scaled) {
    x_use <- (x_seq - mean(occ_data$occ.covs[[focal_var]], na.rm = TRUE)) /
      sd(occ_data$occ.covs[[focal_var]], na.rm = TRUE)
    term_name <- paste0("scale(", focal_var, ")")
  } else {
    x_use <- x_seq
    term_name <- focal_var
  }
  
  curve_df <- do.call(rbind, lapply(sp.names, function(sp) {
    
    b0 <- beta[, paste0("(Intercept)-", sp)]
    b1 <- beta[, paste0(term_name, "-", sp)]
    
    psi_mean <- sapply(x_use, function(x) mean(plogis(b0 + b1 * x)))
    
    data.frame(
      x = x_seq,
      mean = psi_mean,
      species = sp
    )
  }))
  
  curve_df$species <- factor(curve_df$species, levels = sp.names)
  
  ggplot(curve_df, aes(x = x, y = mean, color = species, linetype = species)) +
    geom_line(linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    scale_color_manual(values = grDevices::hcl.colors(length(sp.names), "Dark 3")) +
    scale_linetype_manual(values = rep(c("solid", "dashed", "dotted", "dotdash"),
                                       length.out = length(sp.names))) +
    labs(
      x = xlab_text,
      y = "Occupancy probability",
      color = "Species",
      linetype = "Species"
    ) +
    theme_classic(base_size = 10) +
    theme(
      axis.text.y = element_text(hjust = 1, face = "bold"),
      axis.text.x = element_text(face = "bold"),
      axis.title.x = element_text(face = "bold"),
      axis.title.y = element_text(face = "bold"),
      axis.line = element_line(),
      axis.ticks = element_line(),
      panel.border = element_blank()
    )
}
```

## Make continuous plots

```{r}
p_temp_species   <- make_species_curve("temp_seas", "Temperature seasonality", scaled = TRUE)
p_precip_species <- make_species_curve("precip", "Average annual precipitation (mm)", scaled = TRUE)
p_clay_species   <- make_species_curve("clay", "Percentage clay content", scaled = TRUE)
p_water_species  <- make_species_curve("water_dist", "Distance to water (m)", scaled = TRUE)
p_closed_species <- make_species_curve("closed_prop", "Proportion of closed habitat", scaled = FALSE)
p_rough_species  <- make_species_curve("rough_range", "Terrain roughness", scaled = TRUE)
p_hum_species    <- make_species_curve("log_hum_pop", "Log human population density", scaled = TRUE)
```

## Plots protected area 

```{r}
beta <- as.matrix(top_model$beta.samples)
sp.names <- dimnames(top_model$y)[[1]]

# WDPA levels in the data
wdpa_vals <- sort(unique(occ_data$occ.covs$WDPA))

# Build dataframe of posterior mean occupancy by WDPA category for each species
curve_df <- do.call(rbind, lapply(sp.names, function(sp) {
  
  b0 <- beta[, paste0("(Intercept)-", sp)]
  
  wdpa_means <- sapply(wdpa_vals, function(w) {
    
    # baseline = intercept
    eta <- b0
    
    # add WDPA effect if not the reference category
    wdpa_term <- paste0("as.factor(WDPA)", w, "-", sp)
    
    if (wdpa_term %in% colnames(beta)) {
      eta <- eta + beta[, wdpa_term]
    }
    
    mean(plogis(eta))
  })
  
  data.frame(
    WDPA = factor(wdpa_vals, levels = wdpa_vals),
    mean = wdpa_means,
    species = sp
  )
}))

curve_df$species <- factor(curve_df$species, levels = sp.names)

# Plot
p_wdpa_species <- ggplot(curve_df, aes(x = WDPA, y = mean, color = species, linetype = species, group = species)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = grDevices::hcl.colors(length(sp.names), "Dark 3")) +
  scale_linetype_manual(values = rep(c("solid", "dashed", "dotted", "dotdash"),
                                     length.out = length(sp.names))) +
  labs(
    x = "Protected area category",
    y = "Occupancy probability",
    color = "Species",
    linetype = "Species"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.line = element_line(),
    axis.ticks = element_line(),
    panel.border = element_blank(),
    legend.position = "none"
  )

p_wdpa_species

# Remove legend completly
p_wdpa_species <- p_wdpa_species + guides(color = "none", linetype = "none")
```

## Combine

```{r}
blank <- plot_spacer()

allspecies_plot <-
  (p_closed_species + theme(axis.title.y = element_blank()) |
   p_rough_species  + theme(axis.title.y = element_blank()) |
   p_hum_species    + theme(axis.title.y = element_blank())) /
  (p_wdpa_species   + theme(axis.title.y = element_text()) |
   p_temp_species   + theme(axis.title.y = element_blank()) |
   p_precip_species + theme(axis.title.y = element_blank())) /
  (p_clay_species   + theme(axis.title.y = element_blank()) |
   p_water_species  + theme(axis.title.y = element_blank()) |
   blank) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

## Create table

```{r}
# --- Community + species coefficient table (one row per species) ---

fmt_ci <- function(est, low, high) {
  sprintf("%.2f [%.2f, %.2f]", est, low, high)
}

summarise_draws <- function(x) {
  x <- as.numeric(x)
  c(
    estimate = mean(x, na.rm = TRUE),
    lower_95 = as.numeric(quantile(x, 0.025, na.rm = TRUE)),
    upper_95 = as.numeric(quantile(x, 0.975, na.rm = TRUE))
  )
}

# Community-level samples
beta_comm <- as.matrix(top_model$beta.comm.samples)
comm_terms <- colnames(beta_comm)

comm_tab <- data.frame(
  row = "Community",
  stringsAsFactors = FALSE
)

for (term in comm_terms) {
  s <- summarise_draws(beta_comm[, term])
  comm_tab[[term]] <- fmt_ci(
    as.numeric(s["estimate"]),
    as.numeric(s["lower_95"]),
    as.numeric(s["upper_95"])
  )
}

# Species-level samples
beta_sp <- as.matrix(top_model$beta.samples)
sp.names <- dimnames(top_model$y)[[1]]

species_tabs <- lapply(sp.names, function(sp) {
  out <- data.frame(
    row = sp,
    stringsAsFactors = FALSE
  )
  
  for (term in comm_terms) {
    col_name <- paste0(term, "-", sp)
    if (col_name %in% colnames(beta_sp)) {
      s <- summarise_draws(beta_sp[, col_name])
      out[[term]] <- fmt_ci(
        as.numeric(s["estimate"]),
        as.numeric(s["lower_95"]),
        as.numeric(s["upper_95"])
      )
    } else {
      out[[term]] <- NA_character_
    }
  }
  
  out
})

# Combine community + species into one table
coef_table <- do.call(rbind, c(list(comm_tab), species_tabs))
rownames(coef_table) <- coef_table$row
coef_table$row <- NULL

coef_table
write.csv(coef_table, "coefficient_table.csv", row.names = TRUE)
```

